---
title: "Loans - Final Group Project"
author: "Data Detectives"
date: "2024-11-21"
output: 
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.  Introduction
    a.  Loading the dataset
    b.  Pre-processing the data
2.  Analysis of the Current Scenario
3.  Unsupervised Learning
    a.  Split the data
    b.  Logistic Regression Model
    c.  K-Nearest Neighbor Model
    d.  Artificial Neural Network Model
    e.  Support Vector Machine Model
    f.  Decision Tree Model
    g.  Random Forest Model
    h.  XGBoost Model
    i.  Stack Model
    j.  Comparison of the metrics from all models
4.  Conclusion
    a.  Business Case Conclusion
    b.  Lessons Learned and Improvement Points
    c.  Overall Project and Course Conclusion

------------------------------------------------------------------------

# 1. Introduction: Purpose of the Work

Uncollected loans have historically been one of the primary challenges
facing banks and financial institutions, representing a significant
source of financial loss. While these businesses profit considerably
from interest on loans, they must also bear substantial costs when
borrowers fail to repay their debts. These defaults not only strain
profitability but also impact operational stability and long-term
growth. Addressing this issue has become critical for maintaining
financial health and improving lending strategies.

The objective of this work is to build a predictive model to assist
banks in mitigating losses associated with uncollected loans. By
leveraging historical data, the model aims to predict the likelihood of
loan repayment, enabling better decision-making in loan approvals and
risk management. With a robust predictive system, banks can improve
internal financial planning, reduce exposure to high-risk borrowers, and
optimize resource allocation for loan recovery efforts.

The dataset used in this analysis comes from an American financial
services business, spanning the period from April 2008 to September
2015. Among its many features, the dataset includes a key variable,
`loan_status`, which indicates the repayment status of each loan. Using
this and other attributes, the model will help predict whether loans
will be repaid or defaulted. This tool not only aids in reducing
potential losses but also empowers the company to develop more informed
and effective lending policies.

In summary, this report aims to provide a data-driven solution for a
pressing issue in the banking sector by building a predictive model for
loan analysis. The insights derived from this work will help the
business enhance its financial resilience and support better credit risk
assessment practices. <br>

## a. Loading the dataset

Before staring with the analysis and the financial calculations, we need
to get the data and pre-process it. For organizations sake, all the
libraries were called in this first snippet, just so they are not lost
or repeated in different parts of the code.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# All Libraries used in this file:

library(dplyr) #For pre-processing transformations
library(ggplot2) # For plotting graphics
library(scales) # For one of the graphics scale
library(caret)  # For Confusion Matrix
library(class)  # For KNN
library(neuralnet) # For ANN
library(kernlab) # For SVM
library(C50) # For Decision Tree
library(randomForest) # For Random Forest
library(xgboost) # For Xgboost
```

The following code demonstrates how to get the data. By using the
`stringAsFactors = T` function, the data variables will be read as
factors, avoiding a manual transformation of each, as we will need data
as factors on further steps of this report. The `str()` and the
`summary()` functions were commented due to its length, for organization
purposes.

```{r}
df <- read.csv("final_dataset_loan.csv", stringsAsFactors = T)
#str(df)
#summary(df)
```

<br>

## b. Pre-processing the data

<br>

#### Converting the Target Variable

As mentioned, the target variable is the `loan_status` of each customer,
but as we can see below, there are some of these categories that should
be disregarded considering the goal of prediction if the loan is delayed
or not.

```{r}
levels(df$loan_status)
```

To ensure the analysis is focused and aligned with the objective of
predicting whether loans will be collected or not, certain loan statuses
and categories will be excluded from the dataset. Specifically, loans
categorized as `Does not meet the credit policy. Status: Fully Paid` and
`Does not meet the credit policy. Status: Charged Off` will be removed,
as these do not align with the credit policy criteria under
consideration. Additionally, loans currently `In Grace Period` and those
classified as `Issued` will also be excluded, as they are not actively
in the repayment phase and do not provide relevant information for the
binary prediction task.

By narrowing the dataset to include only loans that meet the credit
policy and are within the payment period, the analysis can focus on
predicting loan repayment outcomes effectively. This approach ensures
the predictive model is built on a dataset representative of the target
scenario, thereby enhancing its reliability and applicability in
assessing whether loans will be collected or not.

The following code deletes the rows in which the column `loan_status` is
one of the levels we decided to drop:

```{r}
df <- df[!df$loan_status %in% c("In Grace Period",
                                "Issued",
                                "Does not meet the credit policy. Status:Charged Off", 
                                "Does not meet the credit policy. Status:Fully Paid"), ]
```

<br>

**Creating the Target Variable:**

With the `loan_status` variable appropriately cleaned, we can now create
a binary target variable, `y`. This variable consolidates all "positive"
loan statuses into a single positive group and all "negative" loan
statuses into a single negative group. By transforming the data into a
binary format, the predictive model becomes more straightforward to
interpret and implement.

This approach aligns with the primary objective of the analysis: to
predict whether a loan will be repaid or not. The binary classification
simplifies the predictive process and provides a clear and actionable
output for decision-making. By focusing on this streamlined prediction,
the model delivers practical insights that support the bank's goal of
efficiently assessing credit risk and optimizing its lending strategy.

```{r}
df$y <- ifelse(df$loan_status %in% c("Charged Off", "Default", "Late (31-120 days)"), 0, 1)
str(df$y)
```

As seen, the code below creates a new column `y` on the dataset. Now the
target variable is a binary, numeric column. <br>

#### Deleting Columns:

**Meaningless Columns:**

Certain columns, such as `id`, `prime_id`, and `url`, are excluded from
the dataset as they hold no predictive power and do not contribute to
the modeling process. These columns are essentially identifiers or
metadata that are irrelevant to the prediction task and could introduce
noise if retained.

In addition to these non-informative columns, the `loan_status` variable
is also removed. Since `y` (the binary target variable) is derived
directly from `loan_status`, the latter is inherently redundant and
would create perfect multicollinearity if included in the analysis.
Retaining it could compromise the integrity of the predictive model by
introducing circular reasoning, where the target variable is predicted
using itself. To ensure the model's reliability and validity,
`loan_status` is excluded from the predictors.

```{r}
df <- df[, !names(df) %in% c('loan_status', 'url', 'id', 'member_id')]
#str(df)
```

<br>

**Future Variables:**

Future variables refer to data that would not be available at the time
of making the prediction, such as information about loan repayment or
behaviors occurring after the loan is issued. Including these variables
can lead to data leakage, where the model uses information it wouldn’t
have in real-world scenarios, resulting in overly optimistic performance
during training but poor generalization to new data. To prevent this,
future variables must be excluded during data pre-processing to ensure
the model is trained only on features realistically available at the
time of prediction, making it both valid and reliable for practical use.

Here we delete those:

```{r}
df <- df[, !names(df) %in% c('grade', 'sub_grade','pymnt_plan', 'funded_amnt_inv')]
#str(df)
```

<br>

#### Dealing with NaNs:

From the structure of the dataframe, we could see that missing values
would appear in two forms: as NaNs (as in columns like `tot_coll_amt`)
but also as " " or "n/a". Before handling them, it is important to set a
standard. The snippet below puts all missing values under the name `NA`
format.

```{r}
df[df == ""] <- NA
df[df == "n/a"] <- NA
data.frame(colSums(is.na(df)))
```

Given the substantial number of predictive columns in the dataset, we
opted to remove any columns with more than 10% missing values. When the
proportion of missing data is too high, imputation methods become less
reliable and may introduce unnecessary bias into the dataset. By
excluding these columns, we aim to preserve the integrity of the
analysis and ensure the predictive model is built on robust and
trustworthy data.

```{r}
threshold <- 0.1
na_percentage <- data.frame((colSums((is.na(df))/nrow(df))))

df <- df[, na_percentage <= threshold]
```

Once we have deleted all the columns that have more than 10% of missing
values, lets see if there are any NaNs remaining:

```{r}
missing_values <- sum(is.na(df))
```

Currently, there are `r missing_values` missing values in the dataset
that need to be addressed. The approach to handling missing values
differs between numerical and categorical variables, so each will be
treated separately to ensure an appropriate and effective imputation
strategy:

```{r}
colSums(is.na(df)) 
```

For the numerical variables, the initial plan was to replace missing
values with the column's median, as this is a common and robust
imputation method. However, this approach can still introduce some
degree of bias into the dataset. Given the large number of observations
available, we determined that the potential risk of bias outweighs the
benefit of retaining rows with missing values. Instead, we opted to
remove rows with missing numerical values, as this approach ensures data
integrity while still leaving a sufficient number of observations to
build reliable and effective predictive models.

```{r, warning=FALSE}
df <- df %>%
  filter(across(where(is.numeric), ~ !is.na(.)))
#str(df)
```

For the categorical variables, rather than deleting rows with missing
values, we opted to create a new category labeled "Unknown." This
approach retains all observations while preserving the dataset's
completeness, ensuring that no potentially valuable information is lost.
By assigning missing values to the "Unknown" category, we can account
for their presence in the analysis without introducing bias or
distortion, allowing the predictive model to handle these cases
effectively.

```{r}
df <- df %>%
  mutate(across(c("emp_title", "emp_length", "title", "last_pymnt_d", "last_credit_pull_d"), ~ ifelse(is.na(.), "Unkown", .)))

# the function above transformed these columns into characters, sowe are just going to transform back to factors
df <- df %>%
  mutate(across(where(is.character), as.factor))
```

To ensure that all missing values have been properly handled, the
following code snippet counts the `NA` values across the entire dataset:

```{r}
final_missing_values <- sum(is.na(df))
cat("Total missing values in the dataset:", final_missing_values, "\n")
```

<br>

#### Addressing factor columns with too many levels:

For models such as Logistic Regression, KNN, ANN, and SVM, the data
needs to be encoded or dummified (a process explained later). This
involves creating a separate column for each level of every categorical
variable. However, some factors in the dataset have an excessively high
number of levels, which can make the models computationally expensive
and inefficient.

To address this, we decided to set these high-cardinality variables to
`NULL`, effectively excluding them from the analysis. While these
variables may have some predictive power, the dataset contains
sufficient remaining predictors to build accurate models without them.
It is worth noting that in scenarios where computational power is not a
limitation, this step could be skipped, and all variables could be
included for a potentially more comprehensive analysis.

```{r}
df$emp_title <- NULL
df$title <- NULL
df$issue_d  <- NULL
df$addr_state <- NULL
df$zip_code <- NULL
df$earliest_cr_line <- NULL
df$last_pymnt_d <- NULL
df$last_credit_pull_d <- NULL
```

<br>

#### Getting a random sample of 10% of the dataset:

After addressing the dataset's issues and ensuring it is properly
treated, we decided to proceed with a random sample of 10% of the rows
from the dataset. This step was introduced to accommodate current
computational limitations, ensuring the modeling process remains
efficient and manageable. However, it is important to note that in
scenarios where computational resources are not a constraint, this step
can be entirely skipped, allowing the full dataset to be utilized for
potentially more robust model training.

Here, we considered creating a balanced random sample, consisting of 50%
random samples from negative cases (defaulted loans) and 50% random
samples from positive cases (repaid loans). While this approach could
potentially improve the performance of models that struggle with
imbalanced datasets, we decided against it. Balancing the dataset
artificially could misrepresent the actual distribution of the target
variable and complicate the model's predictions on new, real-world data,
where the distribution will remain imbalanced. By retaining the original
distribution, we ensure the model's performance and predictions remain
realistic and applicable to the actual business context.

```{r}
sample_size <- 0.1
set.seed(12345)
sample_rows <- sample(1:nrow(df), sample_size*nrow(df))
sample_data <- df[sample_rows, ]
#str(sample_data)
```

<br>

#### Encoding Categorical Variables:

As mentioned earlier, most of the models we will be using cannot handle
categorical columns with more than two levels (i.e., non-binary
categorical variables). To address this, we encode the dataset to
transform these categorical variables into a format that the models can
process effectively.

This encoding process, known as dummification or one-hot encoding,
converts each level of a categorical variable into its own separate
column. For example, the `home_ownership` variable is transformed into
new columns such as `ANY`, `MORTGAGE`, `OWN`, etc., with each row
indicating the presence or absence of a specific level. Importantly, the
encoding process excludes one column for each factor, which prevents
issues with multicollinearity by avoiding redundant information. This
ensures the dataset is both interpretable and compatible with the
predictive models we are implementing.

```{r}
encoded_data <- as.data.frame(model.matrix(~ . -1, sample_data))
#str(encoded_data)
#summary(encoded_data)

# To make the term 36/60 months without " "
colnames(encoded_data) <- make.names(colnames(encoded_data))

# Delete cuz is either all 0 or all 1s
encoded_data$purposeeducational <- NULL
encoded_data$policy_code <- NULL
```

<br>

#### Scaling Numerical Variables: remember to adjust text based on SVM

Finally, the last step of cleaning is scaling our data. Further on in
our analysis, we will build models such as a KNN and a ANN; for these
models, our numeric factors have to be comparable to one another. For
this to happen, they have to be on the same scale, so we create a
separate variable that contains scaled data for the those models. We
utilize min max scaling in this instance, so each numeric variable is
now on a scale of 0-1.

```{r}
minmax <- function(x) {
  (x - min(x)) / (max(x)-min(x))
}
scaled_data <- as.data.frame(lapply(encoded_data, minmax))
```

**Business Note:** The data preparation process outlined above involved
extensive steps to clean and structure the dataset for the upcoming
analysis. While understanding these steps is not essential from a
business perspective, it is important to note that the data has been
thoroughly cleaned and prepared to ensure the analysis is accurate,
reliable, and aligned with the project's objectives.

------------------------------------------------------------------------

# 2. Analysis of the Current Scenario

**Business Note:** Before building the predictive models, this section
will focus on understanding the current scenario of the firm and its
loans. We will conduct some preliminary analyses to evaluate client
behaviors and examine potential correlations with loan statuses. These
insights will provide context for the predictive modeling process and
may highlight key factors influencing loan outcomes.

On this following graph we can see the distribution of the target
variable.

```{r, warning=FALSE, echo=FALSE}

ggplot(df, aes(x = factor(y), fill = factor(y))) +
  geom_bar(width = 0.6, show.legend = FALSE) +
  scale_fill_manual(values = c("firebrick", "steelblue")) +
  labs(
    title = "Loans Paid (1) or Defaulted (0)",
    x = "Loan Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12)
  ) +
  geom_text(stat = 'count', aes(label = scales::percent(..count../sum(..count..), accuracy = 0.1)),
            vjust = -0.5, size = 4)

```

<br>

**Business Note:** The graphic below depicts the bank’s current customer
default distribution with approximately 6% of individuals defaulting on
their loans while 94% do not. Although the percentage of defaulters
appears small, the financial impact on the bank is notably significant.
<br>

```{r, echo=FALSE}
# Some calculations for the graphics: revenues, losses, profit.

# Total expected interest for each loan
df$total_expected_interest <- df$funded_amnt * (df$int_rate/100)


# Calculating Total Revenues - interest fees + late fees (received up to date)
revenue_to_date <- sum(df$total_rec_int, na.rm = TRUE) + 
                    sum(df$total_rec_late_fee, na.rm = TRUE) + 
                    sum(df$recoveries, na.rm = TRUE)

projected_revenue <- sum(
  ifelse(df$y == 1, df$total_expected_interest - df$total_rec_int, 0),
  na.rm = TRUE
)

total_revenue <- revenue_to_date + projected_revenue

# Calculating Total Losses/Expenses - unpaid principal + unpaid interest
total_losses <- sum(
  ifelse(df$y == 0, df$out_prncp + (df$total_expected_interest - df$total_rec_int), 0),
  na.rm = TRUE
)

# Calculating Total Profit
total_profit <- total_revenue - total_losses
```

The following graph aims to provide a clear and straightforward
representation of the company's current scenario. To create this visual,
calculations on revenue and losses were performed, offering an overview
of the financial impact of collected and uncollected loans. This helps
to contextualize the firm's situation before diving into more detailed
analyses.

```{r echo=FALSE}
# Create a data frame for the variables
plot_data <- data.frame(
  Category = c("Total Revenue", "Total Losses", "Total Profit"),
  Amount = c(total_revenue, total_losses, total_profit)
)

# Reorder the Category factor to ensure the desired order in the plot
plot_data$Category <- factor(plot_data$Category, levels = c("Total Revenue", "Total Losses", "Total Profit"))

# Create the bar plot
ggplot(plot_data, aes(x = Category, y = Amount, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "Comparison of Total Revenue, Losses, and Profit",
    x = "Category",
    y = "Amount ($)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12)
  ) +
  scale_fill_manual(values = c("#00AFBB", "#FC4E07", "#E7B800")) +
  geom_text(
    aes(label = scales::dollar_format(big.mark = ".", decimal.mark = ",", prefix = "$")(Amount)),
    vjust = -0.5,
    size = 4
  )
```

**Business Note:** The total revenue from loans stands at
**\$1,650,779,390**. However, the total losses incurred due to defaults
amount to **\$167,030,367**, leaving the bank with a total profit of
**\$1,482,749,023** after accounting for these losses. <br> While
analyzing total amounts provides valuable insights, a percentage-based
graph was also created, where each measurement is expressed as a
percentage of total revenues (with revenues representing 100%). This
approach is particularly useful for comparison, both now and in the
conclusion. Since the number of rows will vary from now to the stacked
model comparison, total figures would not be directly comparable. By
using percentages, we ensure a consistent basis for evaluating
performance and measuring the impact of the predictive models. <br>

```{r, echo=FALSE}
# Calculate percentages relative to Total Revenue
plot_data <- plot_data %>%
  mutate(Percentage = (Amount / Amount[Category == "Total Revenue"]) * 100)

# Create the percentage bar plot
ggplot(plot_data, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "Percentage Breakdown Relative to Total Revenue",
    x = "Category",
    y = "Percentage of Revenue (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12)
  ) +
  scale_fill_manual(values = c("#00AFBB", "#FC4E07", "#E7B800")) +
  geom_text(
    aes(label = paste0(round(Percentage, 2), "%")),
    vjust = -0.5,
    size = 4
  )
```

**Business Note:** While defaulted loans represent only 6% of the total
loans granted, their financial impact is disproportionately high,
contributing to over **10%** of the company's financial losses. This
highlights the significant risk posed by even a small percentage of
defaults.

------------------------------------------------------------------------

# 3. Unsupervised Learning

<br>

Once with the data cleaned and pre-processed, and the analysis of the
current scenario concluded, we are ready to start with the unsupervised
learning. <br>

**Business Note:** This segment focuses on building seven prediction
models—Logistic Regression, K-Nearest Neighbors, Neural Network, Support
Vector Machine, Decision Tree, Random Forest, and XGBoost—to identify
potential loan defaults. The goal is to determine if avoiding high-risk
loans based on predictions can deliver better financial outcomes
compared to the current approach. <br>

## a. Split the data

Data splitting is essential in model training as it enables us to
evaluate a model's performance on data it hasn't encountered before. By
dividing our data into training and test sets, we create two separate
groups: one for training the model (the training set) and another for
evaluating its performance (the test set). This method helps us avoid
overfitting, where the model could learn the training data too well and
struggle to generalize to new, unseen data.

In this setup, we use a test split rate of **0.5**, dividing the dataset
equally into 50% training and 50% testing for each model configuration.
Given the large number of rows in the dataset, allocating half for
training is sufficient to ensure robust model development while leaving
ample data for testing. This approach provides a reliable basis for
evaluating model performance and enables firmer analysis and comparison
of financial results in the end of this report (as the test set is
bigger). Moreover, the 50-50 split reduces computational demands for
heavier models like Artificial Neural Networks (ANN) and Support Vector
Machines (SVM), ensuring the modeling process remains efficient without
compromising accuracy or insights.

Also, we did separate splits for network models vs. the rest of them. We
sample rows from the `scaled_data` dataset to create training and test
sets for the K-Nearest Neighbors (KNN), Artificial Neural Network (ANN),
while for all the other models we are splitting the `encoded_data`
dataset, in which the data was not scaled.

Scaling the dataset is crucial for models like K-Nearest Neighbors (KNN)
and Artificial Neural Networks (ANN) because these algorithms are
sensitive to the magnitude of the input features. In KNN, the distance
between data points determines their classification or prediction, and
unscaled data can give undue weight to features with larger numerical
ranges. Similarly, ANN models rely on gradient-based optimization, where
unscaled inputs can lead to inefficient or biased training due to larger
feature values dominating the gradients. In contrast, models such as
Logistic Regression, Support Vector Machines (SVM), Decision Trees,
Random Forest, and XGBoost do not require scaling. Logistic Regression
can handle unscaled data as it estimates probabilities directly. SVM
uses kernel functions, and tree-based methods (Decision Trees, Random
Forest, XGBoost) split data based on feature thresholds, which are
scale-independent. Therefore, scaling is only necessary for KNN and ANN
to ensure optimal performance, while it can be omitted for the other
models.

```{r}
test_rate <- 0.5
set.seed(12345)

test_rows <- sample(1:nrow(sample_data), test_rate*nrow(sample_data))
train_data <- sample_data[-test_rows, ]
test_data <- sample_data[test_rows, ]

# Log Regression, SVM, Decision Tree, Random Forest - Encoded
train_encoded <- encoded_data[-test_rows, ]
test_encoded <- encoded_data[test_rows, ]

# KNN, ANN Split - Encoded and Scaled
train_scaled <- scaled_data[-test_rows, ]
test_scaled <- scaled_data[test_rows, ]
```

<br>

## b. Logistic Regression Model

Logistic Regression is a statistical model used for binary
classification tasks, where the goal is to predict one of two possible
outcomes (e.g., yes/no, default/no default). It estimates the
probability of an event occurring by modeling the relationship between
one or more independent variables and a binary dependent variable.
Logistic Regression uses a sigmoid function to map predicted values to
probabilities between 0 and 1, and it applies a threshold (commonly 0.5)
to classify observations into one of the two categories. It is widely
used for its simplicity, interpretability, and effectiveness in
understanding the impact of predictor variables on the target outcome.

```{r, cache = TRUE, warning = FALSE}
log_regression <- glm(as.factor(y) ~ ., data = train_encoded, family = binomial)
prediction_lr <- predict(log_regression, test_encoded, type = "response")

#stepwise_lr <- step(log_regression, direction = "both", trace = 0)
#prediction_stepwise <- predict(stepwise_lr, test_encoded, type = "response")

binary_prediction_lr <- ifelse(prediction_lr < .25, 0, 1)

cm_lr <- confusionMatrix(as.factor(binary_prediction_lr),
                as.factor(test_encoded$y),
                positive = "1")
cm_lr
```

The confusion matrix for the Logistic Regression model provides the
following insights:

1.  **Accuracy**: The model achieved an accuracy of **97.56%**, which is
    very high. This indicates that the Logistic Regression model is
    effective at correctly classifying loan repayment outcomes.

2.  **Kappa**: The Kappa statistic is **0.752**, showing substantial
    agreement between predicted and actual loan statuses, beyond what is
    expected by chance. This demonstrates good performance even with
    some degree of class imbalance.

3.  **Sensitivity**: The sensitivity is **99.42%**, meaning the model
    performs exceptionally well in identifying loans that are likely to
    be repaid (positive class). This is crucial for minimizing false
    negatives.

4.  **Specificity**: The specificity is **67.68%**, suggesting that the
    model has moderate success in identifying defaulted loans (negative
    class). While lower than sensitivity, it is still acceptable given
    the overall performance.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.02%**,
    indicating that when the model predicts a loan will be repaid, it is
    correct the vast majority of the time. This is valuable for
    decision-making regarding loan approvals.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **83.55%**, reflecting robust
    overall performance across both classes.

Overall, the Logistic Regression model demonstrates strong performance,
with high accuracy and sensitivity ensuring that loans likely to be
repaid are effectively identified. However, its specificity is
relatively lower, highlighting a need for improvement in detecting
defaulted loans. Despite this, the model remains a reliable tool for
predicting loan repayment outcomes. The stepwise regression approach was
again commented out due to computational demands and the satisfactory
results of the simpler model. <br>

## c. KNN Model

The K-Nearest Neighbors (KNN) algorithm is a non-parametric,
instance-based machine learning model used for classification and
regression tasks. For classification, it works by identifying
the `k` closest data points (neighbors) to a query point in the feature
space and assigning the majority class of those neighbors to the query.
The model relies heavily on the concept of distance (commonly Euclidean)
and is sensitive to feature scaling. KNN is simple, intuitive, and
effective for datasets where the relationship between features and
target labels is not well-defined by a mathematical function.

```{r, cache = TRUE}
y_col <- which(colnames(train_scaled) == "y")
prediction_knn <- knn(train = train_scaled[,-y_col], 
                      test = test_scaled[,-y_col], 
                      cl = train_scaled[,y_col], 
                      k = 1) 
#summary(prediction_knn)

cm_knn <- confusionMatrix(as.factor(prediction_knn), as.factor(test_scaled$y), positive = "1")
cm_knn
```

Before the analysis of the results we wanted to explain the choice of
the `k`. With a highly imbalanced dataset, the majority class dominates,
and using a larger `k` could exacerbate this issue by allowing more
majority-class neighbors to influence the classification decision. By
setting `k = 1`, the model relies solely on the closest neighbor, which
helps avoid amplifying the imbalance effect and allows the model to
focus on the local behavior of individual data points.

The confusion matrix for the KNN model (k=1) provides the following
insights:

1.  **Accuracy**: The model achieved an accuracy of **93.03%**, which is
    good but slightly lower than the Logistic Regression model. This
    indicates that while the KNN model performs well overall, there is
    some room for improvement, particularly in handling the negative
    class.

2.  **Kappa**: The Kappa statistic is **0.2762**, which reflects only
    fair agreement between predictions and actual values. This suggests
    that the model struggles to balance performance across the
    imbalanced classes.

3.  **Sensitivity**: The sensitivity is **97.13%**, demonstrating that
    the model is highly effective at identifying loans that are likely
    to be repaid (positive class). This is critical for ensuring
    reliable predictions for the majority class.

4.  **Specificity**: The specificity is **27.04%**, indicating that the
    model has significant difficulty identifying defaulted loans
    (negative class). This leads to a high number of false positives,
    highlighting the impact of the class imbalance.

5.  **Positive Predictive Value (PPV)**: The PPV is **95.54%**, meaning
    that when the model predicts a loan will be repaid, it is correct
    most of the time. This is beneficial for decision-making on loan
    approvals.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **62.08%**, reflecting moderate
    overall performance, but the model is biased toward the majority
    class.

The KNN model with `k=1` performs well in identifying loans likely to be
repaid, as evidenced by its high sensitivity and PPV. However, it
struggles to correctly classify defaulted loans, with low specificity
and Kappa values. These weaknesses are likely due to the class imbalance
and the use of `k=1`, which makes the model highly sensitive to local
noise in the data. We attempted adjusting `k` to improve tis
specificity, however, it did not work (likely due to the high imbalance
already mentioned). Addressing the class imbalance (e.g., through
resampling techniques) could improve the model’s performance in
detecting the minority class, however we did not try to do it as we aim
to use this regression on the stack model, and we need to maintain the
same rows of train and test sets throughout all models to do so.

## d. ANN Model

An Artificial Neural Network (ANN) is a machine learning model inspired
by the structure and functioning of the human brain. It consists of
layers of interconnected nodes (neurons) that process input data through
weighted connections. ANNs excel at modeling complex, non-linear
relationships in data by learning patterns through backpropagation,
where weights are adjusted iteratively to minimize errors. They are
particularly powerful for tasks like image recognition, natural language
processing, and predictive modeling, offering flexibility and
adaptability for a wide range of problems.

```{r, cache = TRUE}
model_ann <- neuralnet(y ~ ., data = train_scaled, hidden = c(1), stepmax = 1e7, lifesign = "none", threshold = 0.1)
plot(model_ann)

prediction_ann <- predict(model_ann, newdata = test_scaled)
binary_prediction_ann <- ifelse(prediction_ann > 0.25, 1, 0)

cm_ann <- confusionMatrix(as.factor(binary_prediction_ann), as.factor(test_scaled$y))
cm_ann
```

The confusion matrix for the Artificial Neural Network (ANN) model
provides the following insights:

1.  **Accuracy**: The model achieved an accuracy of **98.39%**,
    indicating strong overall performance in classifying loan repayment
    statuses. This high accuracy highlights the ANN's ability to capture
    patterns in the data effectively.

2.  **Kappa**: The Kappa statistic is **0.833**, reflecting strong
    agreement between the model's predictions and the actual outcomes,
    beyond chance. This demonstrates the ANN's reliability even in a
    scenario with class imbalance.

3.  **Sensitivity**: The sensitivity is **72.88%**, showing that the
    model is moderately effective at correctly identifying loans that
    will not be repaid (negative class). While this is a notable
    improvement, there is still room for enhancement in detecting
    defaulted loans.

4.  **Specificity**: The specificity is **99.98%**, meaning the model is
    nearly perfect at identifying loans that will be repaid (positive
    class). This minimizes false positives, which is beneficial for
    business decision-making.

5.  **Positive Predictive Value (PPV)**: The PPV is **99.48%**,
    indicating that when the model predicts a loan will be repaid, it is
    almost always correct. This precision ensures high confidence in
    positive predictions.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **86.43%**, reflecting the model’s
    strong overall performance across both classes.

The ANN model demonstrates exceptional accuracy, specificity, and
precision, making it highly effective for predicting loans likely to be
repaid. However, its sensitivity indicates some difficulty in
identifying defaulted loans. Adjusting hyperparameters, such as
increasing the number of hidden neurons or layers, could potentially
enhance sensitivity, we did not do it due to computer power limitations.
Nonetheless, this ANN configuration provides robust and reliable
predictions with minimal trade-offs in computational efficiency.

## e. SVM Model

A Support Vector Machine (SVM) is a powerful supervised machine learning
model used for classification and regression tasks. It works by finding
the optimal hyperplane that separates data points into distinct classes
with the maximum margin. The margin is the distance between the
hyperplane and the nearest data points from each class, known as support
vectors. SVMs are versatile and can handle linear and non-linear
relationships in data by using kernel functions, which map the input
features into higher-dimensional spaces where classes become more
separable. This flexibility, combined with its robustness to overfitting
in high-dimensional spaces, makes SVMs highly effective for complex
classification problems.

To identify the best-performing kernel for the Support Vector Machine
(SVM) model, I implemented a function, `try_kernels`, that
systematically tests various kernels and evaluates their performance.
This function trains an SVM model using the specified kernel on the
scaled training dataset (`train_scaled`) and generates predictions on
the scaled test dataset (`test_scaled`). For each kernel, the function
calculates and returns three key metrics from the confusion matrix:
accuracy, kappa, and sensitivity. These metrics provide a comprehensive
evaluation of the model's performance, capturing overall correctness,
agreement beyond chance, and the ability to identify positive cases.

The function is applied to a list of kernels (`kernels`), including
"vanilladot", "polydot", "rbfdot", "tanhdot", "anovadot", "laplacedot",
and "besseldot" using `sapply` to iterate over each kernel. This
streamlined approach makes it easy to compare the results and select the
most suitable kernel for the final SVM model based on performance
metrics, ensuring optimal results for the analysis.

```{r,cache=TRUE, warning = FALSE, message = FALSE}

try_kernels <- function(kernel_name) {
  suppressMessages({
    model <- ksvm(as.factor(y) ~ ., data = train_scaled, kernel = kernel_name)
  })
  predictions <- predict(model, test_scaled)
  cm <- confusionMatrix(as.factor(predictions), as.factor(test_scaled$y), positive = "1")
  
  accuracy <- cm$overall["Accuracy"]
  kappa <- cm$overall["Kappa"]
  sensitivity <- cm$byClass["Sensitivity"]
  
  # Return all metrics
  return(c(accuracy = accuracy, kappa = kappa, sensitivity = sensitivity))
}

kernels <- c("vanilladot", "polydot", "rbfdot", "tanhdot", "anovadot", "laplacedot", "besseldot")
sapply(kernels, try_kernels)

suppressMessages({
  model_svm <- ksvm(as.factor(y) ~ ., data = train_scaled, kernel = "vanilladot")
})
prediction_svm <- predict(model_svm, test_scaled)
cm_svm <- confusionMatrix(as.factor(prediction_svm),
                as.factor(test_scaled$y),
                positive="1")
cm_svm

```

Based on the results of the `try_kernels` function, the best kernel to
apply to your SVM model is "vanilladot" (the linear kernel). Here's why:

1.  Accuracy: The linear kernel ("vanilladot") achieved an accuracy
    of 98.29%, which is the highest among all tested kernels (tied with
    "polydot" and slightly better than "anovadot").

2.  Kappa: The Kappa statistic for the linear kernel is 0.8213, also the
    highest among the kernels. This indicates strong agreement between
    predictions and true labels, beyond what is expected by chance, and
    suggests the model handles class imbalance well.

3.  Sensitivity: The sensitivity for "vanilladot" is 99.95%, indicating
    that the model is excellent at identifying the positive class (loans
    likely to be repaid). This is critical in scenarios where minimizing
    missed positive cases is important.

4.  Simplicity and Efficiency: Linear kernels are computationally less
    expensive compared to more complex kernels like "rbfdot" or
    "polydot," making them a more efficient choice when they perform
    equally well or better.

Once with the kernel chosen we can proceed to train and test the model.
The confusion matrix for the Support Vector Machine (SVM) model with a
linear kernel (`vanilladot`) provides the following insights:

1.  **Accuracy**: The model achieved an accuracy of **98.29%**,
    demonstrating excellent overall performance in predicting loan
    repayment statuses. This high accuracy indicates the linear kernel
    is well-suited for this dataset.

2.  **Kappa**: The Kappa statistic is **0.8213**, reflecting strong
    agreement between predictions and actual outcomes, beyond what would
    occur by chance. This highlights the model's reliability in handling
    imbalanced classes.

3.  **Sensitivity**: The sensitivity is **99.95%**, indicating that the
    SVM is nearly perfect at identifying loans that will be repaid
    (positive class).

4.  **Specificity**: The specificity is **71.56%**, meaning the model
    performs moderately well at identifying loans that will default
    (negative class). While specificity is lower than sensitivity, it is
    sufficient given the strong overall performance.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.26%**, showing
    that when the model predicts a loan will be repaid, it is correct
    the vast majority of the time. This high precision is valuable for
    reducing the risk of approving potentially defaulted loans.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **85.75%**, reflecting the model’s
    robustness across both classes.

The SVM model with the linear kernel demonstrates outstanding
performance, particularly in accuracy and sensitivity, making it a
reliable tool for predicting loan repayment outcomes. While the
specificity could be improved to better detect defaulted loans, the
model strikes a good balance between identifying repaid loans and
minimizing false positives. <br>

## f. Decision Tree Model

A Decision Tree is a machine learning model that predicts outcomes by
recursively splitting the dataset into subsets based on feature values.
It starts with a root node representing the entire dataset, splits into
decision nodes based on the most significant features, and ends at leaf
nodes, which represent the final predictions or outcomes. Decision trees
are widely used for classification and regression tasks due to their
simplicity, interpretability, and ability to model complex, non-linear
relationships.

```{r, cache=TRUE}
dt <- C5.0(as.factor(y) ~ ., data = train_encoded)
#plot(dt)

prediction_dt <- predict(dt, newdata = test_encoded)

cm_dt <- confusionMatrix(as.factor(prediction_dt), 
                         as.factor(test_encoded$y), 
                         positive = "1")
cm_dt
```

The confusion matrix for the Decision Tree model provides the following
insights:

1.  **Accuracy**: The model achieved an accuracy of **98.25%**,
    indicating excellent overall performance in predicting loan
    repayment outcomes. This high accuracy reflects the Decision Tree's
    ability to capture meaningful patterns in the dataset.

2.  **Kappa**: The Kappa statistic is **0.817**, signifying strong
    agreement between predicted and actual outcomes beyond chance. This
    highlights the Decision Tree's effectiveness in handling the
    dataset's class imbalance.

3.  **Sensitivity**: The sensitivity is **99.96%**, showing that the
    model is nearly perfect at identifying loans likely to be repaid
    (positive class). This minimizes false negatives, which is critical
    for ensuring reliable predictions of repaid loans.

4.  **Specificity**: The specificity is **70.79%**, indicating moderate
    success in identifying loans that will default (negative class).
    While specificity is lower than sensitivity, it is sufficient given
    the Decision Tree's overall strong performance.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.22%**, meaning
    that when the model predicts a loan will be repaid, it is correct
    the vast majority of the time. This precision is valuable for
    avoiding incorrect approvals of defaulted loans.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **85.37%**, reflecting robust
    overall performance across both classes.

The Decision Tree model demonstrates excellent accuracy, sensitivity,
and precision, making it a reliable tool for predicting loans likely to
be repaid. While the specificity could be improved to better identify
defaulted loans, the model strikes a strong balance between minimizing
false positives and achieving high accuracy. In order to try and improve
the specificity, as it is in our business interest to **reduce false
positives**, we will apply a cost matrix to the model. <br>

### Decision tree with cost matrix

The cost matrix is designed to adjust penalization weights for false
positives and false negatives, prioritizing the reduction of the most
critical error type. In this case, we aim to minimize false
positives—instances where the model predicts a loan will be repaid, but
it defaults. To achieve this, we apply a cost matrix that penalizes
false positives 6 times more than false negatives, aligning the model
with practical financial priorities.

```{r, cache=TRUE}
cost_matrix <- matrix(c(0, 1,
                        6, 0),
                      nrow = 2,
                      byrow = TRUE)

rownames(cost_matrix) <- colnames(cost_matrix) <- c("0", "1")

dt2 <- C5.0(as.factor(y) ~ ., data = train_encoded, costs = cost_matrix)
#plot(dt2)

prediction_dt2 <- predict(dt2, newdata = test_encoded)

cm_dt2 <- confusionMatrix(as.factor(prediction_dt2), as.factor(test_encoded$y), positive = "1")
cm_dt2
```

As a result, most metrics show a slight reduction in performance, though
the model remains robust. However, **specificity has increased**,
aligning with the business goal of minimizing false positives. This
trade-off is reasonable given the purpose of our prediction models. <br>

## g. Random Forest Model

A Random Forest model is an ensemble machine learning algorithm used for
classification and regression tasks. It works by building multiple
decision trees during training and aggregating their predictions to make
a final decision. Each tree is trained on a random subset of the data
and a random selection of features, introducing diversity among the
trees and reducing the risk of overfitting. For classification, the
model predicts the class with the majority vote from all trees, while
for regression, it averages their outputs. Random Forests are highly
effective for handling large datasets, managing missing data, and
capturing non-linear relationships, making them versatile and robust for
various predictive modeling tasks.

```{r, cache=TRUE}

set.seed(12345)
model_rf <- randomForest(as.factor(y) ~ ., data = train_encoded, ntree=100)
#plot(model_rf)

prediction_rf <- predict(model_rf, newdata = test_encoded)

cm_rf <- confusionMatrix(as.factor(prediction_rf),
                as.factor(test_encoded$y),
                positive="1")
cm_rf
```

The confusion matrix for the Random Forest model provides the following
insights:

1.  **Accuracy**: The model achieved an accuracy of **98.26%**,
    reflecting excellent overall performance in predicting loan
    repayment outcomes. This high accuracy demonstrates the Random
    Forest model's capability to capture complex patterns in the data.

2.  **Kappa**: The Kappa statistic is **0.8162**, indicating strong
    agreement between predictions and actual values beyond chance. This
    highlights the model's reliability in handling the dataset's
    imbalanced classes.

3.  **Sensitivity**: The sensitivity is **99.99%**, showing the model is
    nearly perfect at identifying loans that will be repaid (positive
    class). This ensures minimal false negatives, which is crucial for
    reliable classification of repaid loans.

4.  **Specificity**: The specificity is **70.32%**, indicating moderate
    success in identifying loans that will default (negative class).
    While slightly lower, this level of specificity is acceptable given
    the model's focus on minimizing false positives for the majority
    class.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.19%**, meaning
    that when the model predicts a loan will be repaid, it is correct
    the vast majority of the time. This high precision is valuable for
    ensuring confident predictions.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **85.16%**, reflecting robust
    performance across both classes.

The Random Forest model demonstrates excellent accuracy, sensitivity,
and precision, making it a reliable tool for predicting loans likely to
be repaid. While specificity could be improved to better detect
defaulted loans, the model effectively balances its strengths,
especially in identifying and correctly classifying repaid loans. Its
robustness and ability to handle non-linear relationships make it a
strong choice for predicting loan repayment outcomes. Additionally, the
ensemble nature of Random Forest ensures high stability and reliability
in predictions. <br>

## h. XGBoost Model

XGBoost (Extreme Gradient Boosting) is a powerful and scalable machine
learning algorithm based on decision tree ensembles, designed for
classification and regression tasks. It uses a gradient boosting
framework, where models are built sequentially, and each new model
corrects the errors of its predecessor. XGBoost stands out for its
efficiency, regularization techniques to reduce overfitting, and ability
to handle missing data. It supports parallel processing and tree
pruning, making it computationally efficient even with large datasets.
Its adaptability to handle non-linear relationships and high-dimensional
data makes it a top choice for complex predictive modeling tasks, often
delivering superior performance in competitions and real-world
applications.

```{r, cache=TRUE}
set.seed(12345)
labels <- train_encoded[, y_col]

model_xgb <- xgboost(data = as.matrix(train_encoded[, -y_col]),
                    label = as.matrix(labels),
                    nrounds = 100,
                    objective = "binary:logistic",
                    verbose = 0)

prediction_xgb <- predict(model_xgb, as.matrix(test_encoded[, -y_col]))
binary_prediction_xgb <- ifelse(prediction_xgb > 0.5, 1, 0)


cm_xgb <- confusionMatrix(
  as.factor(binary_prediction_xgb),
  as.factor(test_encoded[, y_col]), 
  positive="1"
)
cm_xgb
```

The confusion matrix for the XGBoost model provides the following
insights:

1.  **Accuracy**: The model achieved an accuracy of **98.34%**,
    indicating excellent overall performance in predicting loan
    repayment outcomes. This high accuracy demonstrates the strength of
    XGBoost in handling complex datasets.

2.  **Kappa**: The Kappa statistic is **0.8275**, reflecting strong
    agreement between predictions and actual outcomes beyond chance.
    This indicates that the model handles the class imbalance
    effectively.

3.  **Sensitivity**: The sensitivity is **99.95%**, showing that the
    model is nearly perfect at identifying loans that will be repaid
    (positive class). This ensures a minimal number of false negatives,
    which is critical for ensuring reliable predictions for the majority
    class.

4.  **Specificity**: The specificity is **72.41%**, meaning the model
    performs moderately well at identifying defaulted loans (negative
    class). While lower than sensitivity, it still represents a balanced
    trade-off.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.31%**,
    indicating that when the model predicts a loan will be repaid, it is
    correct the vast majority of the time. This high precision ensures
    confidence in positive predictions.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **86.18%**, reflecting robust
    performance across both classes.

The XGBoost model demonstrates exceptional accuracy, sensitivity, and
precision, making it one of the best-performing models for predicting
loan repayment outcomes. Its specificity, while slightly lower, is still
acceptable given the overall strong performance. XGBoost's ability to
handle class imbalances, non-linear relationships, and feature
importance makes it a highly effective and reliable tool for this
analysis. With its strong predictive power and balanced accuracy,
XGBoost is a top contender for deployment in scenarios requiring robust
loan repayment predictions. <br>

## i. Stacked Model

**Business Note:** After developing and training seven individual
models, we will combine their strengths by creating a stacked model.
This approach leverages the predictions from all the models to build a
final, more robust model that capitalizes on the unique strengths of
each. By combining insights from multiple algorithms, the stacked model
aims to improve prediction accuracy and reliability, providing a
comprehensive solution for identifying loan repayment outcomes. This
method ensures a more confident and data-driven approach to
decision-making.

The first step in building a stacked model is to consolidate all
predictions from the individual models into a single dataset. This
dataset will serve as the input for the stacked model, allowing it to
analyze and learn from the combined outputs of the base models.
Alongside the predictions, it is essential to include the target
variable from the test dataset, as this provides the ground truth needed
for training and evaluating the stacked model. By unifying the
predictions and the target variable, we create a comprehensive dataset
that captures the strengths and weaknesses of each individual model,
enabling the stacked model to make more accurate and reliable
predictions. The code below demonstrates how this process is
implemented:

```{r, cache=TRUE}
response <- test_data$y
all_predictions <- data.frame(lr = binary_prediction_lr, 
                              knn = prediction_knn, 
                              ann = binary_prediction_ann, 
                              svm = prediction_svm, 
                              dt = prediction_dt,
                              rf = prediction_rf,
                              xgb = binary_prediction_xgb, 
                              response)
#str(all_predictions)

# Converting all into factors
all_predictions$lr <- as.factor(all_predictions$lr)
all_predictions$ann <- as.factor(all_predictions$ann)
all_predictions$xgb <- as.factor(all_predictions$xgb)
all_predictions$response <- as.factor(all_predictions$response)
```

Once the new dataset is created, the next step is to split it into
training and testing subsets. This follows the same concept as the
earlier data preparation, where the training subset is used to develop
the stacked model, and the testing subset is reserved for evaluating its
performance. By maintaining this separation, we ensure that the stacked
model is validated on unseen data, providing an unbiased assessment of
its predictive capabilities.

The training subset will include the predictions from the base models
and the target variable, enabling the stacked model to learn how to best
combine the individual predictions. The testing subset will then be used
to evaluate the accuracy and reliability of the stacked model, ensuring
its readiness for practical application. The code below demonstrates
this process:

```{r, cache=TRUE}
test_rate <- 0.5
set.seed(123)

test_prediction_rows <- sample(1:nrow(all_predictions), test_rate*nrow(all_predictions))
train_predictions <- all_predictions[-test_prediction_rows, ]
test_predictions <- all_predictions[test_prediction_rows, ]
```

Now, we proceed to create the stacked model, which is built as a
decision tree applied to the dataset containing all the predictions from
the individual models. The stacked model leverages the combined
predictions of these models to make a final decision, improving overall
predictive performance. By using a decision tree, the stacked model
learns to identify patterns in how the base model predictions interact
and which combinations are most indicative of the target outcome.

The decision tree serves as a meta-model, optimizing the strengths of
the individual models and minimizing their weaknesses. This approach
ensures that the final predictions are more robust and accurate, as they
are derived from the collective insights of all the base models. The
code snippet below illustrates the creation of the stacked model.

```{r,cache=TRUE}
stack_model <- C5.0(as.factor(response) ~ ., data = train_predictions)
#plot(stack_model)

prediction_stack <- predict(stack_model, newdata = test_predictions)

cm_stack <- confusionMatrix(as.factor(prediction_stack),
                            as.factor(test_predictions$response),
                            positive="1")
cm_stack
```

The confusion matrix for the stacked model provides the following
insights:

1.  **Accuracy**: The stacked model achieved an accuracy of **98.33%**,
    demonstrating excellent overall performance in predicting loan
    repayment outcomes. This high accuracy highlights the effectiveness
    of combining multiple model predictions into a single, cohesive
    solution.

2.  **Kappa**: The Kappa statistic is **0.8272**, indicating strong
    agreement between the model's predictions and actual outcomes,
    beyond chance. This suggests that the stacked model handles class
    imbalances effectively.

3.  **Sensitivity**: The sensitivity is **99.97%**, meaning the model is
    almost perfect at identifying loans that will be repaid (positive
    class). This ensures minimal false negatives, a critical factor for
    predicting repaid loans accurately.

4.  **Specificity**: The specificity is **72.16%**, indicating moderate
    performance in identifying defaulted loans (negative class). While
    lower than sensitivity, this is balanced by the model's overall
    strength in classifying the positive class.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.29%**,
    demonstrating that the model is highly reliable when predicting
    loans that will be repaid. This precision is crucial for reducing
    risks associated with misclassified defaulted loans.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **86.06%**, reflecting robust
    overall performance across both classes.

The stacked model effectively combines the strengths of the individual
base models, resulting in exceptional accuracy, sensitivity, and
precision. While specificity could still be improved, the balanced
accuracy indicates that the model performs well across both classes. By
leveraging the insights from all the base models, the stacked model
provides a highly reliable and comprehensive solution for predicting
loan repayment outcomes. This approach ensures informed and confident
decision-making, making it a valuable asset for the business.

To enhance the specificity of our final stacked model, which is crucial
for our business purpose, we will implement a **cost matrix** to assign
greater penalties to false positives. By doing so, the model will
prioritize minimizing the misclassification of defaulted loans (negative
class), thus improving its ability to correctly identify them.

This approach aligns with the strategy used previously for the decision
tree model, where a cost matrix was applied to adjust the model's focus
based on business needs. The cost matrix ensures that the model's
predictions better reflect the practical implications of
misclassifications, especially in scenarios where avoiding false
positives is as important as achieving high overall accuracy.

The updated model with the cost matrix will provide a more balanced
performance, ensuring it meets both operational and business objectives
effectively.

```{r, cache=TRUE}
cost_matrix <- matrix(c(0, 1,   
                       50, 0),  
                       nrow = 2,
                       byrow = TRUE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("0", "1")

stack2_model <- C5.0(as.factor(response) ~ ., data = train_predictions, costs = cost_matrix)

prediction_stack2 <- predict(stack2_model, newdata = test_predictions)

cm_stack2 <- confusionMatrix(as.factor(prediction_stack2),
                             as.factor(test_predictions$response),
                             positive="1")
cm_stack2
```

After applying a cost matrix to penalize false positives more heavily,
the updated stacked model provides the following insights:

1.  **Accuracy**: The model achieved an accuracy of **98.29%**,
    maintaining excellent overall performance. The slight reduction
    compared to the previous stacked model reflects the trade-off
    introduced by prioritizing specificity.

2.  **Kappa**: The Kappa statistic is **0.8253**, indicating strong
    agreement between predictions and actual outcomes, beyond chance.
    This suggests the model continues to handle class imbalance
    effectively, even with the adjusted cost matrix.

3.  **Sensitivity**: The sensitivity is **99.87%**, showing that the
    model remains highly effective at identifying loans that will be
    repaid (positive class). This ensures that very few repaid loans are
    misclassified as defaulted.

4.  **Specificity**: The specificity improved to **73.01%**, reflecting
    a stronger ability to identify defaulted loans (negative class).
    This improvement is directly attributable to the cost matrix, which
    penalizes false positives more heavily.

5.  **Positive Predictive Value (PPV)**: The PPV is **98.34%**,
    demonstrating that the model is still highly reliable when
    predicting loans that will be repaid. This high precision ensures
    confident predictions for the majority class.

6.  **Balanced Accuracy**: The balanced accuracy, which averages
    sensitivity and specificity, is **86.44%**, representing a slight
    improvement and reflecting better performance across both classes.

The introduction of the cost matrix successfully improved the model's
specificity, enhancing its ability to detect defaulted loans while
maintaining strong sensitivity and accuracy. This adjustment aligns the
model more closely with the business goal of minimizing false positives,
ensuring a balanced and effective predictive solution. The updated
stacked model provides greater reliability in identifying defaulted
loans, making it an even more valuable tool for informed
decision-making. <br>

## j. Comparison of the metrics from all models

With all the models built, trained, and tested, we compiled their key
performance metrics — **Accuracy**, **Specificity**, **Sensitivity**,
and **Kappa** — into a single table for comparison. This allows us to
evaluate how each model performs across these metrics and identify their
strengths and weaknesses. Below is the table summarizing the results:

```{r,, echo=FALSE}
metrics <- data.frame(
  Model = c("Logistic Regression", "KNN", "ANN", "Vanilladot SVM", "Decision Tree", "Random Forest", "XG Boost", "Final Stack Model"),
  Accuracy = c(cm_lr$overall["Accuracy"],
               cm_knn$overall["Accuracy"],
               cm_ann$overall["Accuracy"],
               cm_svm$overall["Accuracy"],
               cm_dt2$overall["Accuracy"],
               cm_rf$overall["Accuracy"],
               cm_xgb$overall["Accuracy"],
               cm_stack2$overall["Accuracy"]),
  Specificity = c(cm_lr$byClass["Specificity"],
                  cm_knn$byClass["Specificity"],
                  cm_ann$byClass["Specificity"],
                  cm_svm$byClass["Specificity"],
                  cm_dt2$byClass["Specificity"],
                  cm_rf$byClass["Specificity"],
                  cm_xgb$byClass["Specificity"],
                  cm_stack2$byClass["Specificity"]),
  Sensitivity = c(cm_lr$byClass["Sensitivity"],
                  cm_knn$byClass["Sensitivity"],
                  cm_ann$byClass["Sensitivity"],
                  cm_svm$byClass["Sensitivity"],
                  cm_dt2$byClass["Sensitivity"],
                  cm_rf$byClass["Sensitivity"],
                  cm_xgb$byClass["Sensitivity"],
                  cm_stack2$byClass["Sensitivity"]),
  Kappa = c(cm_lr$overall["Kappa"],
            cm_knn$overall["Kappa"],
            cm_ann$overall["Kappa"],
            cm_svm$overall["Kappa"],
            cm_dt2$overall["Kappa"],
            cm_rf$overall["Kappa"],
            cm_xgb$overall["Kappa"],
            cm_stack2$overall["Kappa"])
)

# Print the metrics table
print(metrics)
```

In the given business scenario, where recognizing false positives
(individuals predicted to repay loans but who actually default) is
critical, Specificity becomes the key metric. This is because
Specificity measures the model's ability to correctly identify the
negative class (defaulted loans), thereby minimizing false positives.

From the table, here are the insights:

1.  Logistic Regression performs well overall but has lower Specificity,
    meaning it struggles to identify defaulted loans, leading to more
    false positives.

2.  KNN has the lowest Specificity, making it unsuitable for this
    business context despite its decent Sensitivity. It would result in
    excessive false positives.

3.  ANN excels in Specificity, nearly perfect at identifying defaulted
    loans, but has low Sensitivity, which might result in overlooking
    some repayable loans.

4.  SVM balances Sensitivity and Specificity, with reasonable
    performance in identifying defaulted loans while maintaining
    excellent accuracy and Sensitivity.

5.  Decision Tree performs relatively well in Specificity, striking a
    balance between recognizing defaulted loans and maintaining good
    overall performance.

6.  Random Forest delivers strong Sensitivity but moderate Specificity,
    making it less ideal for minimizing false positives in this context.

7.  XGBoost strikes a good balance, with improved Specificity and high
    Sensitivity, making it a solid choice for this business scenario.

8.  The Stacked Model, incorporating insights from all models, offers a
    balanced approach. Its improved Specificity and strong Sensitivity
    make it particularly suited for the business objective. <br>

### Recommendation:

For the given business goal of reducing false positives (individuals
predicted to repay loans but who actually default) is critical.
Therefore, Specificity becomes the key metric. This is because
Specificity measures the model's ability to correctly identify the
negative class (defaulted loans), thereby minimizing false positives.

-   ANN is the best at minimizing false positives due to its nearly
    perfect Specificity. However, its low Sensitivity might not
    generalize well across all scenarios.

-   The Final Stacked Model emerges as the best compromise, offering a
    strong balance between Specificity (77.10%) and Sensitivity
    (95.51%). It effectively identifies defaulted loans while
    maintaining overall accuracy and reliability.

Thus, the **Final Stacked Model** is the recommended choice, aligning
well with the business priorities of reducing false positives while
ensuring robust overall performance.

**Business Note:** We have reached the conclusion of the predictive
modeling phase and finalized the stacked model, which integrates the
strengths of all individual models developed in this project. This final
model offers a comprehensive and accurate approach to predicting loan
defaults, providing valuable insights for decision-making. In the next
phase, we will evaluate the financial impact of implementing this model,
focusing on how it can reduce losses from defaulted loans and improve
overall profitability for the company.

------------------------------------------------------------------------

# 4. Conclusion

<br> **Business Note:** In this conclusion segment we want to test if
the model we have built can bring any financial benefits for the
business in reducing the losses on defaulted loans.

To proceed, we need to create a new dataset that includes the rows from
the second split into the test data. This dataset will be augmented with
a new column containing the predictions from the **final stacked
model**. The purpose of this column is to indicate which rows the model
predicts as **positive** (1 = loan repaid). This setup allows us to
clearly identify which loans the model would approve, enabling further
analysis of its financial impact.

```{r}
new_df <- test_data[test_prediction_rows, ]
new_df <- new_df %>%
    mutate(prediction = prediction_stack2)

#str(new_df)
```

To validate that the random sample used for testing retains the overall
distribution of the target variable, we will plot the same distribution
graph as used in the initial analysis. This step ensures that the test
data is representative of the original dataset, providing reliable and
unbiased insights when evaluating the predictions of the final stacked
model.

```{r, echo=FALSE}
ggplot(new_df, aes(x = factor(y), fill = factor(y))) +
  geom_bar(width = 0.6, show.legend = FALSE) +
  scale_fill_manual(values = c("firebrick", "steelblue")) +
  labs(
    title = "Distribution of Target Variable Levels",
    x = "Target Variable",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12)
  ) +
  geom_text(stat = 'count', aes(label = scales::percent(..count../sum(..count..), accuracy = 0.1)),
            vjust = -0.5, size = 4)
```

As expected, the random sample from the test data follows a very
simmilar distribution to the entire dataset.

```{r, echo=FALSE}
# We do the same calculations as before but now on a filtered dataset in which we only consider the rows the final stacked model predicted as 1. 

# Total expected interest for each loan
new_df <- new_df %>%
  filter(prediction == 1)

new_df$total_expected_interest <- new_df$funded_amnt * (new_df$int_rate/100)


# Calculating Total Revenues - interest fees + late fees (received up to date)
new_revenue_to_date <- sum(new_df$total_rec_int, na.rm = TRUE) + 
                       sum(new_df$total_rec_late_fee, na.rm = TRUE) + 
                       sum(new_df$recoveries, na.rm = TRUE)

new_projected_revenue <- sum(
    ifelse(new_df$y == 1, new_df$total_expected_interest - new_df$total_rec_int, 0),
    na.rm = TRUE
)


new_total_revenue <- new_revenue_to_date + new_projected_revenue

# Calculating Total Losses/Expenses - unpaid principal + unpaid interest

new_total_losses <- sum(
  ifelse(new_df$y == 0, new_df$out_prncp + (new_df$total_expected_interest - new_df$total_rec_int), 0),
  na.rm = TRUE
)

new_total_profit <- new_total_revenue - new_total_losses
```

**Business Note:** We will now look at the same financial graphic we
have done for the overall analysis but now with the model applied,
meaning, only offering loans for the people the model predicted would
repaid them. What would be the performance if we only conceded loans to
people the model predict are going to pay back?

```{r, echo=FALSE}
# Create a data frame for the variables
new_plot_data <- data.frame(
  Category = c("Total Revenue", "Total Losses", "Total Profit"),
  Amount = c(new_total_revenue, new_total_losses, new_total_profit)
)

# Reorder the Category factor to ensure the desired order in the plot
new_plot_data$Category <- factor(new_plot_data$Category, levels = c("Total Revenue", "Total Losses", "Total Profit"))

# Create the bar plot
ggplot(new_plot_data, aes(x = Category, y = Amount, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "Comparison of Total Revenue, Losses, and Profit",
    x = "Category",
    y = "Amount ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#00AFBB", "#FC4E07", "#E7B800")) +
  geom_text(
    aes(label = scales::dollar_format(big.mark = ".", decimal.mark = ",", prefix = "$")(Amount)),
    vjust = -0.5,
    size = 4
  )
```

Although evaluating the financial impact using overall values is
insightful, it can be challenging to compare these results with the
current scenario since the total values differ due to the use of
different datasets. As previously explained, the total values are not
directly comparable because the model was applied to a subset of the
data.

To address this, we will create a percentage-based graphic, similar to
the one used earlier. This approach normalizes the results, presenting
each metric (e.g., revenue, losses) as a percentage of total revenue. By
using percentages, we can directly compare the financial outcomes with
and without the model, regardless of the differences in dataset size or
composition.

This percentage-based visualization will provide a clearer understanding
of the relative financial benefits of implementing the model, enabling a
fair and objective comparison with the current scenario.

```{r, echo=FALSE}
# Calculate percentages relative to Total Revenue
new_plot_data <- new_plot_data %>%
  mutate(Percentage = (Amount / Amount[Category == "Total Revenue"]) * 100)

# Create the percentage bar plot
ggplot(new_plot_data, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "Percentage Breakdown Relative to Total Revenue",
    x = "Category",
    y = "Percentage of Revenue (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12)
  ) +
  scale_fill_manual(values = c("#00AFBB", "#FC4E07", "#E7B800")) +
  geom_text(
    aes(label = paste0(round(Percentage, 2), "%")),
    vjust = -0.5,
    size = 4
  )

```

<br>

### a. Business Case Conclusion:

**Before Applying the Model:**

Prior to using the model, total losses accounted for **10.12%** of total
revenue, reflecting a significant portion of loans that resulted in
financial losses. Consequently, total profits represented **89.88%** of
total revenue, indicating profitability but leaving room for improvement
due to the impact of defaulted loans.

**After Applying the Model:**

With the predictive model implemented, total losses decreased to
**7.76%** of total revenue, demonstrating a substantial reduction in
financial losses by identifying and avoiding high-risk loans.
Simultaneously, total profit increased to **92.24%** of total revenue,
highlighting an improvement in profitability as a result of more
informed loan approval decisions.

[**Key Insights:**]{.underline}

1.  **Reduction in Losses**: The model reduced losses by **2.36
    percentage points** (from 10.12% to 7.76%), showcasing its
    effectiveness in minimizing exposure to default risks. While this
    percentage change might not initially appear significant, 2.36% of
    the overall revenue accounts to **3.912,393.604 dollars**.This
    financial impact highlights the meaningful contribution of the
    predictive model in improving the company’s bottom line.

2.  **Increase in Profitability**: Profitability improved by **2.36
    percentage points** (from 89.88% to 92.24%), reflecting the
    financial benefits of prioritizing loans with higher likelihoods of
    repayment.

3.  **Enhanced Risk Management**: The predictive model successfully
    improved the balance between risk and reward by reducing the
    proportion of loans that result in losses while maintaining a high
    level of revenue generation. <br>

The comparison underscores the financial advantages of implementing the
predictive model. By leveraging its ability to filter high-risk loans,
the company achieves a more optimized lending strategy, reducing losses
and enhancing overall profitability. This demonstrates the model's
strategic value in improving the company’s financial outcomes and risk
management practices.

<br>

### b. Lessons Learned and Improvement Points

Throughout this project, we encountered numerous decision points that
involved meaningful trade-offs and required careful consideration. For
example, determining whether to balance the dataset or deciding on the
train/test split ratio were practical challenges that sparked valuable
discussions. These decisions highlighted an essential aspect of data
science: even in a mathematically precise and seemingly objective field,
there are no universally "correct" answers. Instead, each decision
depends on the context, objectives, and constraints of the project.
Reflecting on these trade-offs underscored the importance of aligning
technical choices with real-world implications and business goals.

Regarding areas for improvement, the primary limitation we faced was
computational power. The dataset was exceptionally large, providing an
opportunity to develop highly accurate and complex models. However, the
lack of sufficient computational resources constrained our ability to
explore more sophisticated approaches. For instance, it would have been
valuable to implement a stepwise logistic regression or train a deeper
Artificial Neural Network (ANN) with additional layers and neurons.
These methods could have potentially yielded even better results but
were beyond our computational capacity. <br>

### c. Overall Project and Course Conclusion

The overarching conclusion of this report is that predictive models are
revolutionizing industries and reshaping markets. However, the
effectiveness of these models is intrinsically tied to the quality and
relevance of their input data. As the volume of data being collected
continues to grow exponentially, the potential for predictive models to
become even more accurate and impactful is immense.

In the context of our specific project, we demonstrated how relatively
simple predictive models can generate substantial value for a business
with minimal implementation costs. The potential to increase profits by
nearly 4 million dollars solely by integrating these models into
decision-making processes is a striking outcome. It underscores the
transformative power of data-driven strategies.

Lastly, this project serves as a reminder of the broader potential of
advanced predictive analytics. If such results can be achieved by a team
of university students with limited resources, one can only imagine the
possibilities unlocked by cutting-edge predictive models deployed by
industry leaders with access to high-end computational capabilities.
These tools have the power to shape and transform entire industries,
paving the way for smarter, more efficient, and highly innovative
business strategies.
