---
title: "Final Quiz"
author: "Beatriz Wahle"
date: "2024-11-17"
output: 
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Telecom Data Churn

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# All Libraries used in this file:
library(class)  # For KNN
library(caret)  # For Confusion Matrix
library(ggplot2) # For plotting graphics
library(kernlab) # For SVM
library(C50) # For Decision Trees
library(randomForest) # For Random Forest
```

### Introduction

This Individual Final Project will consist of two main parts, a stacked
model predictive modelling approach and an implementation of the
prediction model. Before getting into an explanation to each, lets have
an overview of the dataset.

**Dataset Overview: Telco Customer Churn**

The dataset analyzed is the Telco Customer Churn dataset, which contains
information on 7,043 customers of a telecommunications company. The
primary objective of this dataset is to predict customer churn, i.e.,
whether a customer will discontinue the service (Churn) based on various
customer demographics, account information, and service usage metrics.
The target variable Churn is a categorical variable with two levels: Yes
(indicating churn) and No (indicating retention).

Regarding the structure of the data: the dataset includes 21 variables,
which are a mix of categorical and numerical data types.

**Part 1: Stacked Model Predictive Modelling Approach**

In this first part of the project we will work with building different
prediction models, for instance a Logistic Regression, a Support Vector
Machine, and a KNN. Although it was aready explained what Churn is, it
is also importat to grasp why it's prediction matter.

[Why Predict Churn?]{.underline}

-   Financial Impact: Churn results in revenue loss for the company, and
    predictive insights can help in designing targeted retention
    campaigns, reducing costs associated with customer acquisition.

-   Competitive Advantage: Minimizing churn ensures better customer
    retention compared to competitors in the telecommunications market.

------------------------------------------------------------------------

### 1. Get the data

The first step before getting into the part 1 of building models is
getting the dataset. This is what I an going to do in the following
snippet. Apart from reading the csv file, the function
`stringsAsFactors = T` will transform the string columns into factors,
making it easier for processes on further steps.

```{r}
cust_data <- read.csv("TelcoChurn.csv", stringsAsFactors = T)
```

With the data loaded, lets look at its summary and structure before any
pre-processing measures.

```{r}
summary(cust_data)
str(cust_data)
```

------------------------------------------------------------------------

### 2. Clean the data

Once we have the data, it is always fundamental to check for missing
values. The code below will count the total of missing values (if any).

```{r}
missing_values <- sum(is.na(cust_data))
```

The dataset has `r missing_values` missing values and we need to handle
those. Let's look at which columns have missing values:

```{r}
data.frame(colSums(is.na(cust_data)))
```

As we can see, the only column with missing values is the
`TotalCharges`. Since it is a numeric column, we could input the mean or
the median on those missing values, however, that can introduce a small
bias. As there are only 11 rows out of the 7043 observations, that
shouldn't justify the exclusoin of the entire column. I will choose to
simply delete those rows.

```{r}
cust_data <- na.omit(cust_data)

# Check there are no missing values
sum(is.na(cust_data))
```

Now that we do not have any missing values, we can move on to the next
steps of pre-processing.

**Deleting meaningless columns:**

The column `customerID` should not aggregate value to the prediction
models and therefore will be deleted.

```{r}
cust_data$customerID <- NULL
```

If we look back to the data structure, we can see that the target column
`Churn` is a factor with two levels. These are coded as 1 and 2 (as
counting in R starts on 1, differently than Python or other languages
that counting starts in 0). Simply for a personal choice of formality
(to maintain the positive class as 1), I will transform the factors into
a numeric column of 0, 1.

```{r}
str(cust_data$Churn)
cust_data$Churn <- ifelse(cust_data$Churn == "No", 0, 1)
```

**Encoding/Creating Dummies:**

I now convert the original dataset into one that represents each
factor/level within a column into it's own separate column
(dummies/one-hot encoding). For example, InternetService turns into
`InternetServiceFiber optic`, and `InternetServiceNo`. One thing to note
is that the model matrix line drops one of the columns for each factor,
as to avoid multicolinearity; this was not the case for the gender
column. Likely due to the inconsistent formatting of the name of the
levels of that variable, our model.matrix line wouldn't drop a column,
so I manually dropped one.

```{r}
encoded_data <- as.data.frame(model.matrix( ~ . - 1, data = cust_data))
encoded_data$genderMale <- NULL

# Just to ensure the structure
str(encoded_data)
```

**Scaling:**

Another important step of pre-processing is scaling the numerical
variables of the data. Further on in this code, I will build a KNN
model; for it, the numeric factors have to be comparable to one another.
For this to happen, they have to be on the same scale, so I create a new
dataframe that contains scaled data for the KNN model. I utilize min-max
scaling in this instance, so each numeric variable is now on a scale of
0-1.

```{r}
# Formual for scaling
minmax <- function(x) {
  (x - min(x)) / (max(x)-min(x))
}

scaled_data <- as.data.frame(lapply(encoded_data, minmax))
```

**Balancing/Oversampling:**

Another possible step before splitting the data and building the models
is balancing the rows in regards to the target variable. On very
imbalanced datasets, some models tend to favor the majority class,
resulting in poor performance on the minority class. So first, lets look
at the current distribution/balance of the Churn column.

```{r}
table(scaled_data$Churn)
positive_percentage <- ((sum(scaled_data$Churn))/(sum(nrow(scaled_data))))*100
```

The dataset has a `r round(positive_percentage, 2)`% proportion of
positive outcomes, indicating a moderate class imbalance. Considering
that one of the primary goals is to build a stacked model, which
requires maintaining the same number of rows across all datasets used
for individual models, this means that either oversampling must be
applied to all models, or none.

While balancing the dataset can enhance the performance of some models,
such as Logistic Regression and K-Nearest Neighbors (KNN), it can
introduce redundancy and negatively impact the performance of others,
such as Support Vector Machines (SVM), Decision Trees, and Random
Forests, particularly when applied to new data. Given these
considerations, and the fact that the level of imbalance is not severe,
I have decided not to balance the dataset to preserve its original
structure and ensure consistency across all models.

------------------------------------------------------------------------

### 3. Split the data

Although some steps of the pre-processing are very important for some of
the models, for some others they are not necessary or should be either
avoided. For that, I will split 3 different dataset, the original one
(but with the customer ID column dropped and no missing values), the
encoded dataset, and the scaled dataset. The models will use the
datasets as following:

1.  Logistic Regression: encoded data. It’s important not to use scaled
    data for Logistic Regression because scaling reduce variance in the
    dataset, which is crucial for Logistic Regression to perform
    accurately.

2.  KNN: scaled data. We should scale data for a KNN model because it
    relies on distance metrics (e.g., Euclidean distance), and unscaled
    features with larger ranges can disproportionately influence the
    model's predictions, leading to biased results.

3.  SVM: scaled data. Support Vector Machine models also need the
    factors to be encoded. Above that you should generally scale your
    data before using a SVM algorithm because they are sensitive to the
    scale of features, meaning features with larger values will have a
    disproportionate influence on the decision boundary, so scaling
    ensures all features contribute equally to the classification
    process.

4.  Decision Tree and Random Forest: both models here should input data
    that is neither encoded or scaled. Both models can handle
    categorical variables directly if they are coded as factors in R and
    are insensitive to the scale of numerical variables.

```{r}
test_rate <- 0.5
set.seed(666)
test_rows <- sample(1:nrow(cust_data), test_rate*nrow(cust_data))
cust_test <- cust_data[test_rows, ]
cust_train <- cust_data[-test_rows, ]

cust_encoded_test <- encoded_data[test_rows, ]
cust_encoded_train <- encoded_data[-test_rows, ]

cust_scaled_test <- scaled_data[test_rows, ]
cust_scaled_train <- scaled_data[-test_rows, ]
```

------------------------------------------------------------------------

### 4. Build models and Predicions:

**Logistic Regression:**

```{r}
lr <- glm(Churn ~ ., data = cust_encoded_train, family = binomial)
stepwise_lr <- step(lr, direction = "both", trace = 0)

prediction_lr <- predict(stepwise_lr, cust_encoded_test, type = "response")
binary_prediction_lr <- ifelse(prediction_lr < .30, 0, 1) # Explain why 0.3 (balance between accuracy/kappa and sensitivity)

confusionMatrix(as.factor(binary_prediction_lr),
                as.factor(cust_encoded_test$Churn),
                positive = "1")
```

**KNN:**

```{r}
evaluate_k <- function(train_data, test_data, train_labels, test_labels, k_values, positive_label) {
  results <- data.frame(
    k = integer(),
    Accuracy = numeric(),
    Kappa = numeric(),
    Sensitivity = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (k in k_values) {
    prediction_knn <- knn(train = train_data, 
                          test = test_data, 
                          cl = train_labels, 
                          k = k)
    
    cm <- confusionMatrix(as.factor(prediction_knn), as.factor(test_labels), positive = positive_label)
    
    results <- rbind(results, data.frame(
      k = k,
      Accuracy = cm$overall["Accuracy"],
      Kappa = cm$overall["Kappa"],
      Sensitivity = cm$byClass["Sensitivity"]
    ))
  }
  
  return(results)
}

k_values <- seq(1, 30, by = 1)

churn_col <- which(colnames(cust_scaled_train) == "Churn")
find_k_train_data <- cust_scaled_train[,-churn_col]
find_k_test_data <- cust_scaled_test[,-churn_col]
find_k_train_labels <- cust_scaled_train[,churn_col]
find_k_test_labels <- cust_scaled_test[,churn_col]


results <- evaluate_k(
  train_data = find_k_train_data,
  test_data = find_k_test_data,
  train_labels = find_k_train_labels,
  test_labels = find_k_test_labels,
  k_values = k_values,
  positive_label = "1"
)

print(results)

ggplot(results, aes(x = k)) +
  geom_line(aes(y = Accuracy), color = "#FF5733") +
  geom_line(aes(y = Kappa), color = "#4CB3FE") +
  geom_line(aes(y = Sensitivity), color = "#6CBB3C") +
  labs(
    title = "KNN Performance Metrics",
    x = "k",
    y = "Metric Value"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))

```

```{r}
churn_col <- which(colnames(cust_scaled_train) == "Churn")
prediction_knn <- knn(train = cust_scaled_train[,-churn_col], 
                      test = cust_scaled_test[,-churn_col], 
                      cl = cust_scaled_train[,churn_col], 
                      k = 7)
summary(prediction_knn)

confusionMatrix(as.factor(prediction_knn), as.factor(cust_scaled_test$Churn), positive = "1")
```

**SVM:**

```{r}
try_kernels <- function(kernel_name) {
  model <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train, kernel = kernel_name)
  predictions <- predict(model, cust_scaled_test)
  cm <- confusionMatrix(as.factor(predictions), as.factor(cust_scaled_test$Churn), positive = "1")
  
  accuracy <- cm$overall["Accuracy"]
  kappa <- cm$overall["Kappa"]
  sensitivity <- cm$byClass["Sensitivity"]
  
  # Return all metrics
  return(c(accuracy = accuracy, kappa = kappa, sensitivity = sensitivity))
}


kernels <- c("vanilladot", "polydot", "rbfdot", "tanhdot", "anovadot", "laplacedot", "besseldot", "splinedot")
data.frame(sapply(kernels, try_kernels))

svm <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train,
                          kernel = "polydot")
svm_prediction <- predict(svm, cust_scaled_test)
confusionMatrix(as.factor(svm_prediction),
                as.factor(cust_scaled_test$Churn),
                positive="1")

```

**Decision Tree:**

```{r}
dt <- C5.0(as.factor(Churn) ~ ., data = cust_train)
plot(dt)

dt_prediction <- predict(dt, newdata = cust_test)

confusionMatrix(as.factor(dt_prediction),
                as.factor(cust_test$Churn),
                positive="1")

cost_matrix <- matrix(c(0, 2,   
                        1, 0),  
                      nrow = 2,
                      byrow = TRUE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("0", "1")


dt2 <- C5.0(as.factor(Churn) ~ ., data = cust_train, costs = cost_matrix)
plot(dt2)

dt2_prediction <- predict(dt, newdata = cust_test)

confusionMatrix(as.factor(dt2_prediction),
                as.factor(cust_test$Churn),
                positive="1")
```

**Random Forest:**

```{r}
set.seed(666)
rf <- randomForest(as.factor(Churn) ~ ., data = cust_train)
plot(rf)
randomForest::varImpPlot(rf)

rf_prediction <- predict(rf, newdata = cust_test)

confusionMatrix(as.factor(rf_prediction),
                as.factor(cust_test$Churn),
                positive="1")
```

------------------------------------------------------------------------

### Gathering all the predictions:

```{r}
# Renaming the original target variable from the test set:
response <- cust_test$Churn

all_predictions <- data.frame(
                              binary_prediction_lr, 
                              prediction_knn, 
                              svm_prediction, 
                              dt2_prediction, 
                              rf_prediction, 
                              response)

str(all_predictions)
all_predictions$binary_prediction_lr <- as.factor(all_predictions$binary_prediction_lr)
all_predictions$response <- as.factor(all_predictions$response)
```

```{r}
test_rate <- 0.3
set.seed(666)

test_prediction_rows <- sample(1:nrow(all_predictions), test_rate*nrow(all_predictions))
train_predictions <- all_predictions[test_prediction_rows, ]
test_predictions <- all_predictions[-test_prediction_rows, ]
```

```{r}
model <- C5.0(as.factor(response) ~ ., data = train_predictions)
plot(model)

predictions <- predict(model, newdata = test_predictions)

confusionMatrix <- table(predictions, test_predictions$response)
print(confusionMatrix)

cost_matrix_prediction <- matrix(c(0, 2,   
                                  1, 0),  
                                  nrow = 2,
                                  byrow = TRUE)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("0", "1")

model2 <- C5.0(as.factor(response) ~ ., data = train_predictions, costs = cost_matrix_prediction)
plot(model2)
predictions2 <- predict(model2, newdata = test_predictions)

confusionMatrix(as.factor(predictions2),
                as.factor(test_predictions$response),
                positive="1")
```
