---
title: "Homework Assignment 1"
author: "Beatriz Wahle"
output:
  html_document:
    toc: yes
    theme: united
    highlight: tango
  pdf_document:
    toc: yes
---

### HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. To complete this assignment, follow these steps:

1.  Download the `HW1.Rmd` file from Canvas.

2.  Open `HW1.Rmd` in RStudio.

3.  Replace the "Your Name Here" text in the `author:` field with names of the students in HW team.

4.  Supply your solutions to the homework by editing `HW1.Rmd`.

5.  When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `HW1_YourNamesHere.Rmd` and then knit again. Submit both the RMD and HTML file on Canvas. (Obv: YourNameHere should be changed to your own name).

6.  Run your code in the Console and Knit HTML frequently to check for errors.

7.  You may find it easier to solve a problem by interacting only with the Console at first.

We will continue working with the nycflights13 dataset we looked at last class.

First - lets read the data into an object named `nyc`.

```{r}
# Please install the nycflights13 package in your console first
library(nycflights13)
nyc <- nycflights13::flights
```

Libraries

```{r}
library(dplyr)
```

I now would like you to answer the following - all in their own separate R code blocks.

### Data Exploration

Let's first do some simple exploration of this data. I would like for you to answer each of the questions below with an **inline R code** integrated into the text and not in a separate R code block.

-   How many airlines are there? (Hint: `levels` and `length` can be useful here)

There are `r length(levels(as.factor(nyc$carrier)))` airlines in the dataset. There are `r length(unique(nyc$carrier, na.rm = TRUE))` airlines in the dataset.

-   How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)

There were `r nrow(nyc[nyc$carrier == 'OO', ])` flights by the airline with code `OO`.

-   How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)

The shortest flight out of NYC is `r min(nyc$air_time, na.rm = TRUE)` minutes.

-   How many flights where there by United Airlines (code: UA) on Jan 12th 2013?

There were `r nrow(nyc[nyc$carrier == 'UA' & nyc$month == 1 & nyc$day == 12,])` flight

There were `r nrow(nyc %>% filter(carrier == 'UA') %>% filter(as.Date(time_hour) == as.Date("2013-01-12")))` flight

(The second filters wrongly some flights from the 11th and misses some of the 12th and I dont understand why! Ask teacher )

### Basics First

You would want to run `str` and `summary` commands and take a look at the output to make sure you have a good understanding of the datatset. Let's answer a few basic questions first.

```{r}
str(nyc)
summary(nyc)
```

-   How many destinations can you fly to?

You can fly to `r length(levels(as.factor(nyc$dest)))` different destinations

You can fly to `r length(unique(nyc$dest, na.rm = TRUE))` different destinations

-   How many flights there were?

There are `r length(levels(as.factor(nyc$flight)))` different flight, but if you are asking for the total number of flights, there were `r nrow(nyc)`.

-   How many airlines are there?

There are `r length(levels(as.factor(nyc$carrier)))` airlines.

-   What is the average distance flown?

The average distance flown is `r mean(nyc$distance, na.rm = TRUE)`

-   What is the median flight time?

The median flight time was `r median(nyc$air_time, na.rm = TRUE)`

-   What is the maximum arrival delay?

The maximum arrival delay was `r max(nyc$arr_delay, na.rm = TRUE)`

### Data Cleaning

Let's do some basic data cleaning.

-   Delete the columns that are not not needed or superfluous.

```{r}
nyc$year <- NULL
nyc$hour <- NULL
nyc$minute <- NULL
```

-   Factorize the columns that should be treated as a factor but is not currently classified as a factor. (Hint: use the `as.factor` command)

```{r}
nyc$carrier <- as.factor(nyc$carrier)
nyc$flight <- as.factor(nyc$flight)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)
```

-   Replace the `NA` values in columns with numerical values with either the mean or the median of that column. Provide explanation for why you used mean or median for replacing `NA` values. (Hint: this will need you to use an ifelse statement. See example below)

```{r}
#Example code for replacing NA values
nyc$arr_delay <- ifelse(is.na(nyc$arr_delay), median(nyc$arr_delay, na.rm = TRUE), nyc$arr_delay)
nyc$dep_delay <- ifelse(is.na(nyc$dep_delay), median(nyc$dep_delay, na.rm = TRUE), nyc$dep_delay)
nyc$air_time <- ifelse(is.na(nyc$air_time), median(nyc$air_time, na.rm = TRUE), nyc$air_time)
nyc$distance <- ifelse(is.na(nyc$distance), median(nyc$distance, na.rm = TRUE), nyc$distance)

# For all of them I opted to use the median in order to respect the distribution of the data. 
# Means can be heavily impacted by outliers, which are possible in all of the columns above
```

### Arrival Delay

Lets focus on Arrival Delay.

-   What was the average arrival delay for all airports and all airlines combined in Jan 2013?

The average arrival delay in Jan 2013 was `r mean(nyc %>% filter(month == 1) %>% pull(arr_delay), na.rm = TRUE)`.

The average arrival delay in Jan 2013 was `r mean(nyc[nyc$month == 1,]$arr_delay, na.rm = TRUE)`.

-   Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?

The average arrival delay in Jan 2013 was `r median(nyc %>% filter(month == 1) %>% pull(arr_delay), na.rm = TRUE)`.

The average arrival delay in Jan 2013 was `r median(nyc[nyc$month == 1,]$arr_delay, na.rm = TRUE)`.

Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

From the differences between the median and the mean we can see the influence of outliers on the dataset. The median being -4 indicates that at least 50% of the flights arrive 4 (minutes?) before time, however there are some flights that might delay a lot from time to time, and those pull the mean up to a delay of 5.8, giving the impression most flights are at least a little late when that\`s not necessarily true.

### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

-   Calculate average arrival delays by airline (Hint: look up the command `tapply`)

```{r}
avr_arrival <- tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)
avr_arrival
```

-   Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)

```{r}
barplot(avr_arrival, las = 2)
```

### Air Gain

Create a new column named `airgain` such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air. (Hint: see below for how to create a new column called Z which is X minus Y)

```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
str(nyc$airgain)
summary(nyc$airgain)
```

Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.

Answer the questions:

-   do airlines actually gain any time when in air on average?

```{r}
mean(nyc$airgain, na.rm = TRUE)
```

Yes, in general airlines gain time on air to make up for eventual delays.

-   Calculate average airgain for different airlines - which airlines do a better job, which do a worse job?

```{r}
airgain_per_airline <- tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)
airgain_per_airline_median <- tapply(nyc$airgain, nyc$carrier, median, na.rm = TRUE)
barplot(airgain_per_airline, las = 2)
barplot(airgain_per_airline_median, las = 2)
```

Evaluating both the mean and the median, the airline that has the best airgain is AS, meanwhile the worse job is from FL. Althought the worse mean is from F9, FL has the 2nd worse mean, and the worst median, probably indicating that they constantly do a worse job.

That's it. Once you are done, make sure everything works and knits well and then you can uplaod the RMD flile and the html output to Canvas.

> Note that I encourage students to collaborate in doing HWs. If you work with someone then please acknowledge that person in your submission. Each student needs to submit his/her file and must have a full understanding of what you are submitting.

**Have Fun!**

Sanjeev
