---
title: "Report - Recommendations to Call Center"
author: "Data Detectives"
date: "2024-10-28"
output: 
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Table of Content

1.  **Introduction**

    a.  Getting and Cleaning the Data

2.  **Descriptive Analysis, Data Visualization and Storytelling**

    a.  Current Scenario

    b.  Break-even Success Rate

    c.  Visualizations

3.  **Supervised Learning**

    a.  Defining the of clusters

    b.  Building the clusters

    c.  Financial Performance/Opportunity

4.  **Unsupervised Learning**

    a.  Balancing the Data

    b.  Splitting the Data

    c.  Logistic Regression Model

    d.  K-Nearest Neighbor Model

    e.  Neural Network Model

    f.  Combined Predictions

    g.  Financial Performance/Opportunity

5.  **Conclusion**

------------------------------------------------------------------------

### Introduction:

This report addresses the current financial challenges faced by the Call
Center and explores potential opportunities through data-driven
solutions. As previously highlighted by Susan Kim and acknowledged by
the entire firm, the business is currently experiencing financial
losses. This report will outline possible strategies for the firm to
achieve break-even or profitability, leveraging data collected over the
recent period.

Given the diverse audience for this report—comprising both business and
data specialists—some sections will include technical documentation.
However, these technical details are structured to ensure that the
primary business goals and opportunities remain clear and accessible at
each stage, leading to actionable financial and business
recommendations.

For data specialists or enthusiasts, the entire report is recommended
reading. For those with a business focus who are primarily interested in
results and recommendations, please refer to sections marked with a
“**Business Note:**” for concise insights and strategic guidance.

#### a. Getting and Cleaning the Data:

As mentioned, this report aims to understand the current situation of
the call center, measure their losses, and find opportunities for it to
break even, or even maybe profit. Before staring with the analysis of
the current situation and the break even calculation, we need to get the
data and clean it. For organizations sake, all the libraries were called
in this first snippet, just so they are not lost or repeated in
different parts of the code.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# All Libraries used in this file:
library(ggplot2)
library(dplyr)
library(factoextra)
library(smotefamily)
library(randomForest)
library(tidyr)
library(caret)
library(neuralnet)
library(class)
```

Before we start with any calculation, it is important to get the data we
will be working with. The following code demonstrates how to do so. By
using the `stringAsFactors = T` function, the data variables will be
read as factors, avoiding a manual transformation of each, as we will
need data as factors on further steps of this report.

```{r}
tele_data <- read.csv("tele.csv", stringsAsFactors = T) 
```

<br>

As aforementioned above, we would like to care of our data prior to
calculations; in particular, cleaning our data. Doing this now will aid
in our analysis, as we don't want any unnecessary columns or improper
data disrupting our analysis.

<br>

**Removing unnecessary columns:**

There are two columns in particular we would like to remove from our
dataset; `X` and `duration`. `X` is just the row count, and is not
pertinent to us at all. `duration` is a post call variable, collected
only after a call has been made. If our goal is to predict future
successful calls, a post call variable is invalid to use, and will thus
be removed; how can we use a variable recorded after a call prior to one
being made?

```{r}
tele_data$X <- NULL
tele_data$duration <- NULL
```

<br> **Pdays column:**

Another column to clean is `pdays`: the number of days that passed by
after the client was last contacted from a previous campaign. Looking at
the distribution of values the column takes, the vast majority of
observations have a value of 999, around 96% to be exact. In this case,
999 represents a client that has never been contacted before. Because of
the distribution and the nature of the column, we converted it into a
binary variable, 0 if a client hasn't been contacted before and 1 if
they have been; this will make the column better formatted to use in our
analysis and easier to interpret. We don't believe we will loose
predictive power by not knowing the time passed from the previous call
from around 3% of the data set.

```{r}
table(tele_data$pdays)
called <- sum(ifelse(tele_data$pdays < 999, 1,0))
never_called <- sum(length(tele_data$pdays))
called_ratio <- (called / (called + never_called))*100
print(paste('The percentage of clients called in this variable is ', round(called_ratio,2),'% of the total database'))
tele_data$pdays <- ifelse(tele_data$pdays == 999, 0, 1)
```

<br> **Encoding:**

We now convert the original dataset into one that represents each
factor/level within a column into it's own separate column
(dummies/one-hot encoding). For example, marital turns into
maritalunknown, maritalsingle, etc. However, one thing to note is that
the model matrix line drops one of the columns for each factor, as to
avoid multicolinearity; this was not the case for the job column. Likely
due to the inconsistent formatting of the name of the levels of that
variable, our model.matrix line wouldn't drop a column, so we manually
dropped one. Additionally, the function changed the name of you
dependent variable to `yyes`, so we converted back to `y`. This is just
a formality to keep the name consistent through all the report.

```{r}
encoded_data <- as.data.frame(model.matrix(~ . -1, tele_data))
encoded_data$jobunknown <- NULL
encoded_data$y <- encoded_data$yyes
encoded_data$yyes <- NULL
```

<br> **Scaling Numerical Variables:**

Finally, the last step of cleaning is scaling our data. Further on in
our analysis, we will build a KNN and ANN model; for these models, our
numeric factors have to be comparable to one another. For this to
happen, they have to be on the same scale, so we create a separate
variable that contains scaled data for the those models. We utilize min
max scaling in this instance, so each numeric variable is now on a scale
of 0-1.

```{r}
minmax <- function(x) {
  (x - min(x)) / (max(x)-min(x))
}
scaled_data <- as.data.frame(lapply(encoded_data, minmax))
```

<br> We have now cleaned our data; removed columns of disinterest,
converted some columns to binary scaling, scaled our data and converted
factors into columns. We are now ready to move on with our analysis.

------------------------------------------------------------------------

### Descriptive Analysis, Data Visualization and Storytelling:

Before trying to find potential solutions, we first want to understand
the problem we are trying to solve and understand our data. We will do
this with a few calculations and visualizations.

#### a. Current Scenario

We were given some financial details about the call center at the start
of the case; using this information, we calculate the current financial
state of the company given the data.

```{r}
successful_calls <- sum(tele_data$y == "yes")
total_calls <- nrow(tele_data)
success_rate <- successful_calls / total_calls

# Define the financial parameters - given by MK - to calculate the BEP
training_cost <- 1000 
variable_cost_per_call <- 1
contribution_margin <- 10
avg_retention <- 1000

total_var_cost <- (total_calls*variable_cost_per_call) + 
                  ((total_calls/avg_retention)*training_cost)

revenue_per_call <- (total_var_cost/total_calls) + contribution_margin
total_revenue <- (success_rate*total_calls) * revenue_per_call

break_even_revenue <- (total_var_cost)/(contribution_margin)
gross_margin <- total_revenue - total_var_cost
```

<br> **Business Note:** Interesting insights we can gather from a simple
data analysis are the following:

1.  The current success rate from calls is **`r success_rate * 100`%**,
    which means that out of the `r total_calls` only
    `r successful_calls` are successful. As already known by the entire
    staff, this is very low.
2.  Since there are no major fixed costs to account for - as mentioned
    by MK - we can only calculate the break-even revenue (no fixed costs
    to deduct). This tells us how much total revenue the company needs
    to generate to cover their variable costs. The current break-even
    revenue for the call center is \$`r break_even_revenue`.
3.  Currently, the call center's gross margin is **$ -26696**. This just confirms what everyone already knew, that the call center is losing money.

In order to solve this problem, instead of simply increasing the number
of total calls, and hoping to increase the total number of successful
calls with it, the more efficient solution is to focus on increasing the
success rate to be able to break even, or even profit, at the current
level of sales. With that in mind, the first question to answer is by
how much would the success rate would have to increase for them to break
even?

In order to answer this question, there is an important detail to take
into account. Besides just considering the current variable costs the
company has, and the contribution margin per call, we have to remember
MK's insight that: "every 1% increase in call success rate increased the
average retention rate by 100 calls."

Getting back to the technical content, we built the following function
to determine what the break-even success rate should be; what portion of
calls needed to be successful for the company to break-even at the
current number of total calls. This calculation also factors in the
increment in average retention rate with a 1% increase in success rate.
While revenue is less than cost (or profit \< 0), the function still
runs, incrementing the success rate by 1% and retention by 100, than
recalculating until we break even. We than get as the return the success
rate upon completion of the while loop.

```{r}
find_success_rate <- function(total_calls, contribution_margin, total_var_cost,  variable_cost_per_call, avg_retention, training_cost, success_rate) {

  while (((success_rate * total_calls)*(contribution_margin+variable_cost_per_call))-total_var_cost < 0) {
    success_rate <- success_rate + 0.01 # increment success rate by 1%
    avg_retention <- avg_retention+100
    total_var_cost <- (total_calls*variable_cost_per_call) + 
                  ((total_calls/(avg_retention))*training_cost) # adjust effective calls
  }
  
  return(success_rate)
}

necessary_success_rate <- find_success_rate(total_calls, contribution_margin, total_var_cost,  variable_cost_per_call, avg_retention, training_cost, success_rate)
```

<br>

**Business Note:** The success rate needed to break-even at the current
level of total calls it **`r necessary_success_rate * 100`%**, which
represent a `r  (necessary_success_rate - success_rate) * 100` %
increase from the current success rate. This imply a total of
`r round((necessary_success_rate * total_calls), digits = 2)` successful
calls per month.

Now, to conclude this first part of the analytics and visualizations, we
are going to display some graphics that will help with a better
understanding of the problem, and potential opportunities of solution.

From Figure 1, displayed below, we can simple compare the current
success rate with the necessary one.

```{r, echo=FALSE}
success_data <- data.frame(
  metric = c("Current Success Rate", "Necessary Success Rate"),
  rate = c(success_rate, necessary_success_rate)
)

ggplot(success_data, aes(x = metric, y = rate)) +
  geom_bar(stat = "identity", fill = "#72D57D") +
  labs(title = "Figure 1: Current vs. Necessary Success Rate", y = "Success Rate", x = "Metric")
```

**Business Note:** Although from this graphic it may seem like there is
a need for a huge improvement, when looking at the overall numbers, we
can see that is a relative small improvement if looking at the total
amount of calls. Look for the green increase from left to right on
Figure 2.

<br>

```{r, echo=FALSE}
failure_rate <- 1 - success_rate
necessary_failure_rate <- 1 - necessary_success_rate

call_outcomes <- data.frame(
  scenario = rep(c("Current", "Necessary"), each = 2),
  outcome = rep(c("Success", "Failed"), 2),
  rate = c(success_rate, failure_rate, necessary_success_rate, necessary_failure_rate)
)

ggplot(call_outcomes, aes(x = scenario, y = rate, fill = outcome)) +
  geom_bar(stat = "identity", position = "stack", width = 0.5) +
  labs(title = "Figure 2: Comparison of Current vs Necessary Success/Failure Rates",
       x = "Scenario", 
       y = "Proportion",
       fill = "Outcome") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Success" = "#72D57D", "Failed" = "#FA746C")) +
  theme_minimal()
```

**Business Note:** This figure is similar to figure 1, except now we
show the successes in relation to failures in both the current and
necessary success rate. This further illustrates how small the
improvement has to be in order for the call center to turn things
around.

<br> Once we have outlined the overall objective to break even. It is
important to understand how to reach that goal. How can the call center
increase the success rate? The answer can seem quite simple, by focusing
the calls on the people that are most likely to subscribe to the term
deposit. To facilitate the visualization, let's look at how the behavior
of customers change in some of the variables offered by the dataset.

```{r, echo=FALSE}
# FIGURE 3: Overall success and failure of calls per age + the success rate per age.

# Create a new data set to organize the data as we need it to plot.
outcomes_per_age <- tele_data %>%
  filter(age <= 80) %>% # from 80 on is too little data an the success rate is meaningless
  group_by(age) %>% # from now on everything is done per each age group
  summarise( # new summary data frame with the following variables/calculations for each group
    total_calls = n(), # total calls per age group
    successful_calls = sum(y == "yes"), # successful calls per age
    failed_calls = sum(y == "no"), # failed calls per age
    success_rate = successful_calls / total_calls # success rate per age
  ) %>% # reshape successful and failed calls into 2 col, one tells you if succ of fail, the other is a count 
  gather(key = "outcome", value = "count", successful_calls, failed_calls)

# For later, to scale the success rate plot line
max_count <- max(outcomes_per_age$count)


ggplot(outcomes_per_age, aes(x = age)) +
  # bar plot for number of successful and failed calls
  geom_bar(aes(y = count, fill = outcome), stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("failed_calls" = "#FFC086", "successful_calls" = "#B7FABF"), 
                    name = "Call Outcome", 
                    labels = c("Failed", "Successful")) +
  # line plot for success rate
  geom_line(aes(y = success_rate * max_count, group = 1), color = "#4E85B5", linewidth = 0.35) +
  geom_point(aes(y = success_rate * max_count), color = "#4E85B5", size = 0.5) +
  scale_y_continuous(
    name = "Number of Calls",
    sec.axis = sec_axis(~ . / max_count, name = "Success Rate", labels = scales::percent)
  ) +
  labs(title = "Figure 3: Number of Successful and Failed Calls by Age with Success Rate",
       x = "Age") +
  theme_minimal()
```

**Business Note:** We also wanted to take a look at how the success rate
varies in relation to certain columns; in this case, Age. We show the
portion of successes compared to failures, the success rate, and the
number of observations for the range of ages. Although they have less
calls when compared to 30-40 year olds, 18-25 year olds and 60-80 year
olds have the highest success rate; this could prove insightful further
into our analysis.

```{r, echo=FALSE}
# FIGURE 4: Overall success and failure of calls per job + the success rate per job

# same as before, but now grouping by job
outcomes_per_job <- tele_data %>%
  group_by(job) %>% 
  summarise(
    total_calls = n(), 
    successful_calls = sum(y == "yes"), 
    failed_calls = sum(y == "no"), 
    success_rate = successful_calls / total_calls 
  ) %>% 
  gather(key = "outcome", value = "count", successful_calls, failed_calls)

max_count <- max(outcomes_per_age$count)

ggplot(outcomes_per_job, aes(x = job)) +
  
  geom_bar(aes(y = count, fill = outcome), stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("failed_calls" = "#FFC086", "successful_calls" = "#B7FABF"), 
                    name = "Call Outcome", 
                    labels = c("Failed", "Successful")) +
  
  geom_line(aes(y = success_rate * max_count, group = 1), color = "#4E85B5", linewidth = 0.35) +
  geom_point(aes(y = success_rate * max_count), color = "#4E85B5", size = 0.5) +
  scale_y_continuous(
    name = "Number of Calls",
    sec.axis = sec_axis(~ . / max_count, name = "Success Rate", labels = scales::percent)
  ) +
  labs(title = "Figure 4: Number of Successful and Failed Calls by Job with Success Rate",
       x = "Job") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Business Note:** Similar to figure 3, we now showcase successes,
failures, and success rate by occupation type; we see that some
occupation types, like admin, blue-collar, and technician, have a high
number of calls, as well as that some occupation types like retired and
student have higher success rates.

```{r, echo=FALSE}
# FIGURE 5: Overall success and failure of calls per campaign + the success rate per campaign

# same as before, but now grouping by campaign
outcomes_per_campaign <- tele_data %>%
  filter(campaign <= 15) %>%
  group_by(campaign) %>% 
  summarise(
    total_calls = n(), 
    successful_calls = sum(y == "yes"), 
    failed_calls = sum(y == "no"), 
    success_rate = successful_calls / total_calls 
  ) %>% 
  gather(key = "outcome", value = "count", successful_calls, failed_calls)

max_count <- max(outcomes_per_campaign$count)

ggplot(outcomes_per_campaign, aes(x = campaign)) +
  
  geom_bar(aes(y = count, fill = outcome), stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("failed_calls" = "#FFC086", "successful_calls" = "#B7FABF"), 
                    name = "Call Outcome", 
                    labels = c("Failed", "Successful")) +
  
  geom_line(aes(y = success_rate * max_count, group = 1), color = "#4E85B5", linewidth = 0.35) +
  geom_point(aes(y = success_rate * max_count), color = "#4E85B5", size = 0.5) +
  scale_y_continuous(
    name = "Number of Calls",
    sec.axis = sec_axis(~ . / max_count, name = "Success Rate", labels = scales::percent)
  ) +
  labs(title = "Figure 5: Number of Successful and Failed Calls by Number of Contacts performed during this Campaign with Success Rate",
       x = "Campaign") +
  theme_minimal()
```

**Business Note:** Finally, we look at the three metrics from before,
except now by the number of calls made to a client. We can see that the
success rate dwindles upon repeated calls, something to note going
forward.

<br> **Business Note:** Upon inspecting our visuals, we have a few
takeaways. We don't need to increase the success rate by too much in
order or the call center to break even. Additionally, the variables of
Age, Job, and Campaign may be of significance when conducting our
analysis; we see clear values for all three that increase the success
rate. We will see if these findings come to fruition, or if they weren't
significant enough.

------------------------------------------------------------------------

### Supervised Learning:

**Business Note:** With a comprehensive understanding of the current
situation and the call center’s distance from break-even, we can begin
exploring opportunities to leverage collected data to improve the
success rate. In this section, we will focus on identifying distinct
clusters of customers—groups of individuals who share similar
characteristics or behaviors. The purpose of clustering is to, once
these groups are defined, analyze the varying success rates across them.
This approach enables associates to prioritize calls to clusters with
higher success rates, optimizing efficiency and potentially boosting
overall performance.

<br>

#### a. Defining the of clusters

The code below is getting the scaled data, deleting the prediction
variable (as you don't want it to be part of the variables used to build
the clusters), and attributing that to a new data, `clustering_data`.
Before constructing clusters, we first need to determine the optimal
number of clusters. This can be done using several techniques, including
Within-Cluster Sum of Squares (WSS), Gap Statistics, and Silhouette
analysis—all of which are built-in functions in R.

However, with a dataset of over 50,000 observations across 53 variables,
these techniques can be computationally intensive and reached our
computers' memory limitations. To address this without increasing memory
allocation on our device, we use Principal Component Analysis (PCA) to
reduce the dimensionality of the data.

PCA transforms the original variables into a smaller set of new,
uncorrelated variables called principal components. Each principal
component captures a certain amount of the total variance in the data.
By selecting only the components that explain the majority of the
variance (for example, 90-95%), we effectively condense the information
from many variables into a few components. This reduces the dataset's
dimensionality, making computations faster and more memory-efficient
without sacrificing significant information.

After applying PCA, we select a subset of components that explains 90%
of the data's variance. This reduced dataset, `clustering_data_pca`, was
still computationally heavy, so we decided to sample 8,000 observations
and from that apply the mentioned techniques (WSS, Gap Statistics, and
Silhouette) to determine the optimal number of clusters.

```{r, cache = TRUE, warning = FALSE, message = FALSE}
y_col <- which(colnames(tele_data) == "y")

clustering_data <- scaled_data[,-y_col]
clustering_data <- as.data.frame(clustering_data)

numeric_data <- clustering_data[sapply(clustering_data, is.numeric)]
pca_result <- prcomp(numeric_data, scale. = TRUE)

explained_variance <- summary(pca_result)$importance[3,]
num_components <- which(cumsum(explained_variance) >= 0.90)[1]  
clustering_data_pca <- pca_result$x[, 1:num_components]

set.seed(666)
clustering_data_pca_sample <- clustering_data_pca[sample(nrow(clustering_data_pca), 8000), ]

fviz_nbclust(clustering_data_pca_sample, kmeans, method = "wss")
fviz_nbclust(clustering_data_pca_sample, kmeans, method = "gap_stat")
fviz_nbclust(clustering_data_pca_sample, kmeans, method = "silhouette")

```

Choosing **k=6** based on the provided plots can be supported with
statistical reasoning, as follows:

-   Within-Cluster Sum of Squares (WSS): The WSS plot shows a decreasing
    trend in total WSS as k increases. There is a noticeable "elbow" or
    reduction in the rate of decrease around k=6, indicating that adding
    more clusters beyond this point yields diminishing returns in terms
    of reducing within-cluster variance. Choosing k=6 thus balances
    between cluster compactness and parsimony, capturing distinct
    customer segments without unnecessary complexity.

-   Gap Statistic: The Gap Statistic plot shows that the optimal k might
    be around 5, as the gap value peaks there. However, the difference
    between k=5 and k=6 is minor, suggesting that k=6 is also a strong
    candidate for maximizing the separation between clusters. Since
    lower values of k do not capture the success rate effectively, k=6
    is an appropriate choice to maintain interpretability while
    capturing meaningful patterns.

-   Silhouette: The Silhouette plot shows a local peak at k=2,
    suggesting high cohesion and separation for very small clusters.
    However, for more granular clustering, k=6 still maintains a
    reasonable silhouette width without a significant drop in quality.
    This implies that the clusters are relatively well-defined and
    distinct at k=6, supporting the choice.

<br>

#### b. Building the clusters

After determining that we will use 6 clusters, we proceed with building
them. The first few lines of code create the clusters using the k-means
algorithm with k=6 and add a new column to `clustering_data` that
specifies the assigned cluster for each row.

Next, we use the `fviz_cluster` function from the `factoextra` library
to visualize the clusters. The first plot shows each data point within
its respective cluster, while the second plot adds ellipses representing
the normal distribution for each cluster, providing a clearer view of
the cluster boundaries and dispersion.

Finally, we integrate the cluster assignments into the original
`scaled_data`, which includes the target variable `y.` Using `tapply`,
we calculate the mean value of `y` for each cluster. This calculation
helps us understand the average outcome or behavior (as represented by
`y`) within each cluster.

```{r, cache=TRUE}
#set.seed(123) 
set.seed(666)
km <- kmeans(clustering_data, 6)
clustering_data$cluster <- km$cluster

fviz_cluster(km, data = clustering_data, geom = "point")
fviz_cluster(km, data = clustering_data, geom = "point", ellipse = TRUE, ellipse.type = "norm", ellipse.level = 0.95, ellipse.alpha = 0.1)

scaled_data$cluster <- km$cluster
tapply(scaled_data$y, scaled_data$cluster,mean)
```

Although the graphics appear to show overlapped clusters, it is
important to remember this is a 2D graphic representing
multi-dimensions, so they are not really overlapped. As for the `y` mean
per cluster, it allows us to identify clusters with higher success rates
(the ones with a higher means) and target them more effectively.

As seen by the output of the `tapply` function, cluster 3 is the one
with the highest mean at 23.6%, followed by cluster 2 with a mean of
20.4%.

**Business Note:** Now that we have the clusters built, we can see which
are likely to be the most profitable ones when looking at the clusters
mean on failed (0) and successful (1) calls. Clusters **2** and **3**
are likely to be the ones with the highest success rate. In the next
segment we would look at how can we convert this concept into convincing
financial figures.

<br>

#### c. Financial Performance/Opportunity

In this segment we come back to the initial calculations of Part A, of
understanding the current scenario, but simulating would would have been
the performance if the call center only called cluster 3, or called
cluster 2 and 3. The other clusters have too low of a mean (lower than
the current success rate) that it is not reasonable to call them as
well.

```{r}
cluster_three <- scaled_data[scaled_data$cluster == 2, ]
c3_total_calls <- nrow(cluster_three)
c3_succesful_calls <- nrow(cluster_three[cluster_three$y==1, ])
c3_success_rate <- c3_succesful_calls/c3_total_calls

c3_revenue <- (c3_success_rate*c3_total_calls) * revenue_per_call

difference_success_rate <- (c3_success_rate - success_rate)*100
c3_total_cost <- (c3_total_calls*variable_cost_per_call) + 
                  ((c3_total_calls/(avg_retention+(100*difference_success_rate)))*training_cost)

c3_gross_margin <- round((c3_revenue - c3_total_cost), 2)
```

```{r}
cluster_three_and_two <- scaled_data[scaled_data$cluster %in% c(2, 3), ]
c32_total_calls <- nrow(cluster_three_and_two)
c32_succesful_calls <- nrow(cluster_three_and_two[cluster_three_and_two$y==1, ])
c32_success_rate <- c32_succesful_calls/c32_total_calls

c32_revenue <- (c32_success_rate*c32_total_calls) * revenue_per_call

difference_success_rate2 <- (c32_success_rate - success_rate)*100
c32_total_cost <- (c32_total_calls*variable_cost_per_call) + 
                  ((c32_total_calls/(avg_retention+(100*difference_success_rate2)))*training_cost)

c32_gross_margin <- round((c32_revenue - c32_total_cost), 2)
```

**Business Note:** After seeing which clusters were likely to be more
profitable, we simulated how the performance of the call center would
have been if the associates have only called cluster 3 (the best
performing one), or called both clusters 2 and 3, and here are the main
takeaways.

-   By only calling cluster 3 (`r c3_success_rate`% success rate), the
    call center would have made a total of `r c3_total_calls` that would
    have generated a profit of \$`r c3_gross_margin`. Although this
    already looks better than the current scenario, there is still room
    for improvement.
-   By calling both clusters 2 and 3 (overall success rate of
    `r c32_success_rate`%), the call center would have called
    `r c32_total_calls` people, making $18204.95
    of profit.

We could see that filtering of the calls made to the clusters with the
highest success rates is a strategy that can boost the call centers
performance. With a `r difference_success_rate2`% increase in the
success rate, making the call center not only break-even, but also
profit.

Although at this point of the report it may seem like we have already
reached a solution to the problem, there is a second approach that is
worth trying and testing the performance, to in the end, choose the
alternative in which the call center can profit the most

------------------------------------------------------------------------

### Unsupervised Learning:

**Business Note:** In this segment of the report, our goal is to build 3
prediction models - a Logistic Regression, a K-Nearest Neighbors and a
Neural Network - and see if by having a prediction of who are the
potential successful calls, and targeting those, deliver a result as
good as (or better than) the previous segment

<br>

#### a. Balancing the Data

Before proceeding, we need to remove the cluster column that was added
to the scaled dataset in the previous segmentation phase. This column is
irrelevant for the prediction model and could introduce noise or bias
into the training process.

Before proceeding, we need to remove the cluster column that was added
to the scaled dataset in the previous segmentation phase. This column is
irrelevant for the prediction model and could introduce noise or bias
into the training process.

Another key point to note is why data balancing is essential and how it
impacts prediction models. When data is unbalanced, meaning one class is
represented significantly more than another, prediction models tend to
be biased towards the majority class. This can lead to poor
generalization and unreliable predictions, especially for the minority
class. By balancing the dataset, we ensure that our models have a fair
representation of each class, enabling them to learn patterns more
effectively from both classes and ultimately making the model more
accurate and generalizable.

We are addressing the class imbalance issue at this stage, rather than
during initial data cleaning, because balancing is a model-specific
preprocessing step.

We will create two balanced datasets: one from the scaled data for
training the K-Nearest Neighbors (KNN) and Artificial Neural Network
(ANN) models, and another using the encoded but unbalanced data for the
Logistic Regression model. It’s important not to use scaled data for
Logistic Regression because scaling reduce variance in the dataset,
which is crucial for Logistic Regression to perform accurately. This
approach allows each model to operate with data that maximizes its
performance potential while preserving statistical integrity.

```{r}
scaled_data$cluster <- NULL

table(scaled_data$y)

smote_data <- SMOTE(scaled_data[, -which(names(scaled_data) == "y")], scaled_data$y, K = 5, dup_size = 2)
balanced_data <- data.frame(smote_data$data)
names(balanced_data)[ncol(balanced_data)] <- "y"
balanced_data$y <- as.numeric(balanced_data$y)

 
smote_data_lr <- SMOTE(encoded_data[, -which(names(encoded_data) == "y")], encoded_data$y, K = 5, dup_size = 2)
balanced_data_lr <- data.frame(smote_data_lr$data)
names(balanced_data_lr)[ncol(balanced_data_lr)] <- "y"
balanced_data_lr$y <- as.numeric(balanced_data_lr$y)

```

To be more specific, our code applies Synthetic Minority Over-sampling
Technique (SMOTE) to balance two datasets, `scaled_data` and
`encoded_data`, by increasing instances of the minority class in the
target variable `y`. For each dataset, it removes y from the features,
applies SMOTE with 5 nearest neighbors and a duplication size of 2, and
converts the result into a new balanced data frame. The target variable
`y` is then renamed and converted to a numeric type in each balanced
dataset, producing `balanced_data` and `balanced_data_lr` as the
balanced versions of `scaled_data` and `encoded_data`.

<br>

#### b. Splitting the Data

Data splitting is essential in model training as it enables us to
evaluate a model's performance on data it hasn't encountered before. By
dividing our data into training and test sets, we create two separate
groups: one for training the model (the training set) and another for
evaluating its performance (the test set). This method helps us avoid
overfitting, where the model could learn the training data too well and
struggle to generalize to new, unseen data.

In this setup, we use a test rate of 0.2 to split our data into 80%
training and 20% testing for each model configuration:

Separate data split for Network Models and Logistic Regression: We
sample rows from the first balanced dataset to create training and test
sets for the K-Nearest Neighbors (KNN), Artificial Neural Network (ANN),
while for the Logistic Regression we are splitting the second balanced
dataset, in which the data was not scaled.

```{r}
test_rate <- 0.2
#set.seed(666)
test_rows <- sample(1:nrow(balanced_data), test_rate*nrow(balanced_data))
train_data <- balanced_data[test_rows, ]
test_data <- balanced_data[-test_rows, ]

test_rows_lr <- sample(1:nrow(balanced_data_lr), test_rate*nrow(balanced_data_lr))
train_data_lr <- balanced_data_lr[test_rows_lr, ]
test_data_lr <- balanced_data_lr[-test_rows_lr, ]
```

<br>

#### c. Logistic Regression Model

In order to optimize our profits and only target customers who are
predicted to buy our product, our team first developed a logistic
regression. This model identifies relationships between various inputs
and predicts the probability of an outcome. Furthermore, rather than
pursuing a computationally-intensive process of building a giant model,
and then a step one that simplifies it, this approach builds a simple
model and then it performs a stepwise selection on the simple model.
Stepwise selection is a method of refining a model by adding or removing
predictors based on specific criteria.

As we want for the prediction to be binary, we create an `ifelse` in
which every prediction below 0.25 is considered to be a failed call (0),
and from 0.25 up a successful call (1).

```{r, cache=TRUE, warning = FALSE, message = FALSE}
lr <- glm(y ~ ., data = train_data_lr, family = binomial)
stepwise_lr <- step(lr, direction = "both", trace = 0)

prediction_lr <- predict(stepwise_lr, test_data_lr, type = "response")
binary_prediction_lr <- ifelse(prediction_lr < .25, 0, 1)

confusionMatrix(as.factor(binary_prediction_lr),
                as.factor(test_data_lr$y),
                positive = "1")
```

In order to analyze the performance of this model, we ran a confusion
matrix. The logistic regression model shows an accuracy of 77.47%,
meaning it correctly classifies around 77% of cases. Its sensitivity
(65.98%) indicates it identifies about 66% of actual positive cases, and
the specificity is higher at 81.84%, showing that the model is better at
identifying true negatives.

<br>  

#### d. K-Nearest Neighbor Model

Next, our team developed a K-Nearest Neighbors model, another
alternative predictive model. KNN creates predictions by examining a
data point’s k nearest neighbors (k being a number we decide upon, 8). A
data point’s nearest neighbors would be ones that are most similar to it
based on the predictive variables.

```{r, cache = TRUE}
y_col <- which(colnames(train_data) == "y")
prediction_knn <- knn(train = train_data[,-y_col], 
                      test = test_data[,-y_col], 
                      cl = train_data[,y_col], 
                      k = 4)
summary(prediction_knn)

confusionMatrix(as.factor(prediction_knn), as.factor(test_data$y), positive = "1")
```

The k-nearest neighbors (KNN) model achieved an accuracy of 77.8%,
correctly classifying around 78% of cases. The specificity (85.42%) is
higher than the sensitivity (57.99%), indicating the model is better at
identifying true negatives than true positives. Overall this model
performs well, but worse than the logistic regression.

<br>

####  e. Neural Network Model

For our third and final prediction model, we built an Artificial Neural
Network (ANN). This model generates predictions by recognizing patterns
within complex data. An ANN consists of individual units called neurons,
which function similarly to linear regressions, connected through
multiple layers. The connections between neurons have weights, which are
adjusted through a process called backpropagation. During
backpropagation, the network iteratively updates these weights to
minimize prediction errors, allowing the model to learn and improve its
accuracy over time.

```{r, cache=TRUE}
ann <- neuralnet(y ~ ., data = train_data, hidden = c(1, 1), stepmax = 1e6, lifesign = "none", learningrate = 0.1)
plot(ann)

prediction_nn <- predict(ann, newdata = test_data)
summary(prediction_nn)

binary_prediction_nn <- ifelse(prediction_nn > 0.2, 1, 0)
confusionMatrix(as.factor(binary_prediction_nn),
                as.factor(test_data$y),
                positive = "1")
```

The neural network model achieved an accuracy of 74.81%, meaning it
correctly classifies about 75% of cases. The sensitivity (70.17%)
indicates that the model successfully identifies around 70% of actual
positive cases, while specificity (76.59%) shows it correctly identifies
approximately 77% of negatives. Overall, the model has fairly balanced
sensitivity and specificity, performing moderately well in identifying
both positive and negative cases.

<br>

#### f. Combined Predictions

To create a more robust prediction model, I combined the outputs of the
three individual models—Logistic Regression, K-Nearest Neighbors (KNN),
and Artificial Neural Network (ANN). Here, we combined the outputs of
the three models—Logistic Regression, K-Nearest Neighbors (KNN), and
Artificial Neural Network (ANN)—into a single prediction using a
weighted voting approach. In this setup, ew assigned double weight to
the ANN model, as it has proven to be the most robust among the three
models. The combined prediction (combined_pred) uses an ifelse condition
to check if the sum of votes (with ANN votes counted twice) is at least
2; if so, it assigns a "1" (positive) prediction, otherwise "0"
(negative). This final combined prediction was then evaluated using a
confusion matrix (combined_confusion_matrix) to assess the performance
of this ensemble model.

```{r}
binary_prediction_lr <- as.factor(binary_prediction_lr)
prediction_knn <- as.factor(prediction_knn)
binary_prediction_nn <- as.factor(binary_prediction_nn)

# Apply weighted voting (example: logistic gets more weight)
combined_pred <- as.factor(ifelse(rowSums(cbind((binary_prediction_lr == "1"), (prediction_knn == "1"), (binary_prediction_nn == "1")*2)) >= 2, "1", "0"))

combined_confusion_matrix <- confusionMatrix(combined_pred, as.factor(test_data$y), positive = "1")
combined_confusion_matrix
```

The combined model achieved an accuracy of 76.75%, meaning it correctly
classified about 77% of cases. The sensitivity (73.78%) indicates that
it successfully identifies around 74% of actual positive cases, while
specificity (77.89%) shows a similar performance in correctly
identifying negatives. Overall, this ensemble model demonstrates a solid
performance across both positive and negative cases.

An important thing to notice is that, although the double weight on the
ANN will decrease the overall predictions success rate in the next step
(decreases from a 61% to a 56%), this model captures more clients, and
make more calls, making it more profitable. The trade-off is that by
decreasing a bit the success rate, we will increase the total amount of
calls to be made, which also increases the gross margin.

<br>

#### g. Financial Performance/Opportunity

**Business Note:** Now, with all the models built and a combined
prediction that leverages each model's strength, we can measure how
successful would be to implement this method for the Call Center. We
will be running similar calculations than for the clustering segment, to
compare the performances.

```{r}
true_positives <- combined_confusion_matrix$table[2, 2]
prediction_false_positives <- combined_confusion_matrix$table[2, 1]
total_positives <- prediction_false_positives+true_positives

prediction_sucess_rate <- (true_positives)/total_positives
prediction_percentage_of_total_calls <- total_positives/nrow(test_data)
prediction_total_calls <- round((prediction_percentage_of_total_calls * total_calls),0)
prediction_successful_calls <- round((prediction_sucess_rate*prediction_total_calls),0)

predition_revenue <- prediction_total_calls * revenue_per_call

difference_success_rate3 <- (prediction_sucess_rate - success_rate)*100
prediction_total_costs <- (prediction_total_calls*variable_cost_per_call) + 
                  ((prediction_total_calls/(avg_retention+(100*difference_success_rate3)))*training_cost)

prediction_gross_margin <- round((predition_revenue - prediction_total_costs), 2)
```

**Business Note:** From the calculations above we can see that if the
call center decided to implement the prediction models technique, they
would have a success rate of `r prediction_sucess_rate`%. Better than
the rate reached by the clustering strategy.

However, what actually matters is how much profit can this method bring.
The combined prediction models would have advised the associate to call 15023 clients, compared to the actual `r total_calls` calls made during that period. This, considering the model's success rate, would imply a profit of $162519.6 to the company.

------------------------------------------------------------------------

### Conclusion:

After thoroughly analyzing the problem and evaluating two potential
strategies for enhancing the Call Center's performance—customer
clustering and predictive modeling—it is clear that both approaches
offer significant improvements given the current business context.
However, the gross margin associated with the predictive modeling
strategy is notably higher. Additionally, there is substantial room for
further refinement within the predictive models, including optimizing
feature selection, exploring more complex models, implementing
cross-validation, and employing advanced ensemble methods. These
improvements could lead to even more substantial financial gains.

In light of these insights, the final recommendation for Susan and her
team is to prioritize the predictive modeling approach in the upcoming
period. By using these models to identify which clients to contact, the
team can increase the success rate, lower operational costs, and
ultimately enhance the firm’s profitability.
