---
title: "IndividualFinalProject_BeatrizWahle"
author: "Beatriz Wahle"
date: "2024-11-21"
output: 
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# All Libraries used in this file:
library(class)  # For KNN
library(caret)  # For Confusion Matrix
library(ggplot2) # For plotting graphics
library(kernlab) # For SVM
library(C50)
```

### Introduction

This Individual Final Project will consist of two main parts, a stacked
model predictive modelling approach and an implementation of the
prediction model. Before getting into an explanation to each, lets have
an overview of the dataset.

**Dataset Overview: Telco Customer Churn**

The dataset analyzed is the Telco Customer Churn dataset, which contains
information on 7,043 customers of a telecommunications company. The
primary objective of this dataset is to predict customer churn, i.e.,
whether a customer will discontinue the service (Churn) based on various
customer demographics, account information, and service usage metrics.
The target variable Churn is a categorical variable with two levels: Yes
(indicating churn) and No (indicating retention).

Regarding the structure of the data: the dataset includes 21 variables,
which are a mix of categorical and numerical data types.

**Part 1: Stacked Model Predictive Modelling Approach**

In this first part of the project we will work with building different
prediction models, for instance a Logistic Regression, a Support Vector
Machine, and a KNN. Although it was already explained what Churn is, it
is also important to grasp why it's prediction matter.

[Why Predict Churn?]{.underline}

-   Financial Impact: Churn results in revenue loss for the company, and
    predictive insights can help in designing targeted retention
    campaigns, reducing costs associated with customer acquisition.

-   Competitive Advantage: Minimizing churn ensures better customer
    retention compared to competitors in the telecommunications market.

[Potential Users of the Predictive Model:]{.underline}

1.  Marketing Teams: To design retention strategies such as targeted
    offers or loyalty programs for customers at high risk of churning.
2.  Customer Support Departments: To proactively engage with at-risk
    customers and resolve their issues before they decide to leave.
3.  Product Managers: To identify patterns in customer preferences and
    dissatisfaction, enabling them to improve products and services.
4.  Senior Management/Executives: To evaluate the effectiveness of churn
    mitigation strategies and allocate resources more effectively.
5.  Data Scientists/Analysts: To continuously improve and refine the
    predictive models, ensuring sustained performance as customer
    behaviors evolve.

------------------------------------------------------------------------

### 1. Get the data

The first step before getting into the part 1 of building models is
getting the dataset. This is what I an going to do in the following
snippet. Apart from reading the csv file, the function
`stringsAsFactors = T` will transform the string columns into factors,
making it easier for processes on further steps.

```{r}
cust_data <- read.csv("TelcoChurn.csv", stringsAsFactors = T)
```

With the data loaded, lets look at its summary and structure before any
pre-processing measures.

```{r}
summary(cust_data)
str(cust_data)
```

------------------------------------------------------------------------

### 2. Clean the data

Once we have the data, it is always fundamental to check for missing
values. The code below will count the total of missing values (if any).

```{r}
missing_values <- sum(is.na(cust_data))
```

The dataset has `r missing_values` missing values and we need to handle
those. Let's look at which columns have missing values:

```{r}
data.frame(colSums(is.na(cust_data)))
```

As we can see, the only column with missing values is the
`TotalCharges`. Since it is a numeric column, we could input the mean or
the median on those missing values, however, that can introduce a small
bias. As there are only 11 rows out of the 7043 observations, that
shouldn't justify the exclusion of the entire column. I will choose to
simply delete those rows.

```{r}
cust_data <- na.omit(cust_data)

# Check there are no missing values
sum(is.na(cust_data))
```

Now that we do not have any missing values, we can move on to the next
steps of pre-processing.

**Deleting meaningless columns:**

The column `customerID` should not aggregate value to the prediction
models and therefore will be deleted.

```{r}
cust_data$customerID <- NULL
```

If we look back to the data structure, we can see that the target column
`Churn` is a factor with two levels. These are coded as 1 and 2 (as
counting in R starts on 1, differently than Python or other languages
that counting starts in 0). Simply for a personal choice of formality
(to maintain the positive class as 1), I will transform the factors into
a numeric column of 0, 1.

```{r}
str(cust_data$Churn)
cust_data$Churn <- ifelse(cust_data$Churn == "No", 0, 1)
```

**Encoding/Creating Dummies:**

I now convert the original dataset into one that represents each
factor/level within a column into it's own separate column
(dummies/one-hot encoding). For example, `InternetService` turns into
`InternetServiceFiber optic`, and `InternetServiceNo`. One thing to note
is that the model matrix line drops one of the columns for each factor,
as to avoid multicolinearity; this was not the case for the gender
column. Likely due to the inconsistent formatting of the name of the
levels of that variable, our model.matrix line wouldn't drop a column,
so I manually dropped one.

```{r}
encoded_data <- as.data.frame(model.matrix( ~ . - 1, data = cust_data))
encoded_data$genderMale <- NULL

# Just to ensure the structure
str(encoded_data)
```

**Scaling:**

Another important step of pre-processing is scaling the numerical
variables of the data. Further on in this code, I will build a KNN
model; for it, the numeric factors have to be comparable to one another.
For this to happen, they have to be on the same scale, so I create a new
dataframe that contains scaled data for the KNN model. I utilize min-max
scaling in this instance, so each numeric variable is now on a scale of
0-1.

```{r}
# Formual for scaling
minmax <- function(x) {
  (x - min(x)) / (max(x)-min(x))
}

scaled_data <- as.data.frame(lapply(encoded_data, minmax))
```

**Balancing/Oversampling:**

Another possible step before splitting the data and building the models
is balancing the rows in regards to the target variable. On very
imbalanced datasets, some models tend to favor the majority class,
resulting in poor performance on the minority class. So first, lets look
at the current distribution/balance of the Churn column.

```{r}
table(scaled_data$Churn)
positive_percentage <- ((sum(scaled_data$Churn))/(sum(nrow(scaled_data))))*100
```

The dataset has a `r round(positive_percentage, 2)`% proportion of
positive outcomes, indicating a moderate class imbalance. Considering
that one of the primary goals is to build a stacked model, which
requires maintaining the same number of rows across all datasets used
for individual models, this means that either oversampling must be
applied to all models, or none.

While balancing the dataset can enhance the performance of some models,
such as Logistic Regression and K-Nearest Neighbors (KNN), it can
introduce redundancy and negatively impact the performance of others,
such as Support Vector Machines (SVM), Decision Trees, and Random
Forests, particularly when applied to new data. Given these
considerations, and the fact that the level of imbalance is not severe,
I have decided not to balance the dataset to preserve its original
structure and ensure consistency across all models.

------------------------------------------------------------------------

### 3. Split the data

The final step before getting into the part of building the models (Part
1) is splitting the dataset into train and test.

Although some steps of the pre-processing are very important for some of
the models, for some others they are not necessary or should be either
avoided. For that, I will split 3 different dataset, the original one
(but with the customer ID column dropped and no missing values), the
encoded dataset, and the scaled dataset. The models will use the
datasets as following:

1.  Logistic Regression: encoded data. It’s important not to use scaled
    data for Logistic Regression because scaling reduce variance in the
    dataset, which is crucial for Logistic Regression to perform
    accurately.

2.  KNN: scaled data. We should scale data for a KNN model because it
    relies on distance metrics (e.g., Euclidean distance), and unscaled
    features with larger ranges can disproportionately influence the
    model's predictions, leading to biased results.

3.  SVM: scaled data. Support Vector Machine models also need the
    factors to be encoded. Above that you should generally scale your
    data before using a SVM algorithm because they are sensitive to the
    scale of features, meaning features with larger values will have a
    disproportionate influence on the decision boundary, so scaling
    ensures all features contribute equally to the classification
    process.

```{r}
test_rate <- 0.5
set.seed(666)
test_rows <- sample(1:nrow(cust_data), test_rate*nrow(cust_data))

cust_encoded_test <- encoded_data[test_rows, ]
cust_encoded_train <- encoded_data[-test_rows, ]

cust_scaled_test <- scaled_data[test_rows, ]
cust_scaled_train <- scaled_data[-test_rows, ]
```

------------------------------------------------------------------------

## Part 1: Stacked Model Predictive Modelling Approach

**Logistic Regression:**

This model identifies relationships between various inputs and predicts
the probability of an outcome. Furthermore, rather than pursuing a
computationally-intensive process of building a giant model, and then a
step one that simplifies it, this approach builds a simple model and
then it performs a stepwise selection on the simple model. Stepwise
selection is a method of refining a model by adding or removing
predictors based on specific criteria.

As I want for the prediction to be binary, I created an `ifelse` in
which every prediction below 0.3 is considered to be failed (0), and
from 0.3 up successful (1). The threshold was decided to be 0.3 because
after some manual iterations (e.g. 0.2, 0.25, 0.4, 0.5), this was the
one to better balance the metrics of accuracy, sensitivity and
specificity.

```{r}
lr <- glm(Churn ~ ., data = cust_encoded_train, family = binomial)
stepwise_lr <- step(lr, direction = "both", trace = 0)

prediction_lr <- predict(stepwise_lr, cust_encoded_test, type = "response")
binary_prediction_lr <- ifelse(prediction_lr < 0.3, 0, 1) 

confusionMatrix(as.factor(binary_prediction_lr),
                as.factor(cust_encoded_test$Churn),
                positive = "1")
```

The logistic regression model achieves an accuracy of 75.82%, which is
good, with a moderate Kappa value of 0.4497, indicating fair agreement
beyond chance. Sensitivity, at 75.30%, shows the model effectively
identifies most churners but still misses about a quarter, which could
impact retention strategies.

**Support Vector Machine:**

A Support Vector Machine (SVM) predicts by finding the optimal
hyper-plane or boundary that maximally separates data points of
different classes in a high-dimensional space, using support vectors
(critical data points) to define this margin.

Here, we were requested to test out 3 different kernels, and find out
which performs the best for our dataset. There are two ways to do so:
(1) manually testing out all of the kernels or (2) having a function
that loops over the kernels and gather the most important values to
compare them (Accuracy, Kappa and Sensitivity).

Although both codes are demonstrated below, I chose to use the function
simply because the visualization of all the metrics of the 3 kernels
together makes it easier for comparison.

```{r}
set.seed(666)
try_kernels <- function(kernel_name) {
  model <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train, kernel = kernel_name)
  predictions <- predict(model, cust_scaled_test)
  cm <- confusionMatrix(as.factor(predictions), as.factor(cust_scaled_test$Churn), positive = "1")
  
  accuracy <- cm$overall["Accuracy"]
  kappa <- cm$overall["Kappa"]
  sensitivity <- cm$byClass["Sensitivity"]
  
  # Return all metrics
  return(c(accuracy = accuracy, kappa = kappa, sensitivity = sensitivity))
}


kernels <- c("vanilladot", "polydot", "rbfdot")
data.frame(sapply(kernels, try_kernels))



svm <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train,
                          kernel = "polydot")
svm_prediction <- predict(svm, cust_scaled_test)
confusionMatrix(as.factor(svm_prediction),
                as.factor(cust_scaled_test$Churn),
                positive="1")

#svm2 <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train,
#                          kernel = "vanilladot")
#svm_prediction2 <- predict(svm2, cust_scaled_test)
#confusionMatrix(as.factor(svm_prediction2),
#                as.factor(cust_scaled_test$Churn),
#                positive="1")
#
#svm3 <- ksvm(as.factor(Churn) ~ ., data = cust_scaled_train,
#                          kernel = "rbfdot")
#svm_prediction3 <- predict(svm3, cust_scaled_test)
#confusionMatrix(as.factor(svm_prediction3),
#                as.factor(cust_scaled_test$Churn),
#                positive="1")
```

Based on the information provided we can conclude that:

-   Accuracy: All three kernels (vanilla, polynomial, and RBF) achieve
    nearly identical accuracy (\~79.8%), meaning they perform similarly
    in predicting both churners and non-churners.

-   Kappa: The vanilla and polynomial kernels have slightly higher Kappa
    values (0.4551) compared to the RBF kernel (0.4333). This indicates
    that the vanilla and polynomial kernels better align predictions
    with the actual outcomes beyond random chance.

-   Sensitivity: The vanilla and polynomial kernels show the highest
    sensitivity (0.5541), meaning they are better at identifying
    churners than the RBF kernel.

Conclusion: The **vanilla** kernel or **polynomial** kernel best fits
the data due to their higher Kappa and sensitivity, making them more
effective for churn prediction.

Regarding the performance of the best model chosen, the SVM model using
the "polydot" kernel achieves an accuracy of 79.81%, which is a
noticeable improvement over the Logistic Regression. The Kappa statistic
(0.4551) reflects moderate agreement beyond random chance, comparable to
other models evaluated. However, the sensitivity (55.41%), still leaves
room for improvement in identifying churners effectively. The high
specificity (88.39%) shows the model is better at predicting
non-churners, but a more balanced sensitivity would enhance its
effectiveness for churn prediction tasks.

**K-Nearest Neighbors:**

Next, I developed a K-Nearest Neighbors model, another predictive model.
KNN creates predictions by examining a data point’s k nearest neighbors
(k being a number we decide upon). A data point’s nearest neighbors
would be ones that are most similar to it based on the predictive
variables.

In order to choose the k in which the model performs best, I developed a
function that will run through ks from 1 to 20 to give me each Accuracy,
Kappa, and Sensitivity. Then, I will plot it so I can look for the
optimal K. (The code here is hidden due to its length)

```{r, echo=FALSE}
evaluate_k <- function(train_data, test_data, train_labels, test_labels, k_values, positive_label) {
  results <- data.frame(
    k = integer(),
    Accuracy = numeric(),
    Kappa = numeric(),
    Sensitivity = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (k in k_values) {
    prediction_knn <- knn(train = train_data, 
                          test = test_data, 
                          cl = train_labels, 
                          k = k)
    
    cm <- confusionMatrix(as.factor(prediction_knn), as.factor(test_labels), positive = positive_label)
    
    results <- rbind(results, data.frame(
      k = k,
      Accuracy = cm$overall["Accuracy"],
      Kappa = cm$overall["Kappa"],
      Sensitivity = cm$byClass["Sensitivity"]
    ))
  }
  
  return(results)
}

k_values <- seq(1, 20, by = 1)

churn_col <- which(colnames(cust_scaled_train) == "Churn")
find_k_train_data <- cust_scaled_train[,-churn_col]
find_k_test_data <- cust_scaled_test[,-churn_col]
find_k_train_labels <- cust_scaled_train[,churn_col]
find_k_test_labels <- cust_scaled_test[,churn_col]


results <- evaluate_k(
  train_data = find_k_train_data,
  test_data = find_k_test_data,
  train_labels = find_k_train_labels,
  test_labels = find_k_test_labels,
  k_values = k_values,
  positive_label = "1"
)

print(results)

ggplot(results, aes(x = k)) +
  geom_line(aes(y = Accuracy), color = "#FF5733") +
  geom_line(aes(y = Kappa), color = "#4CB3FE") +
  geom_line(aes(y = Sensitivity), color = "#6CBB3C") +
  labs(
    title = "KNN Performance Metrics",
    x = "k",
    y = "Metric Value"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))

```

Based both on the table and the plot (although might be harder to
pin-point at it), the k that maximizes the 3 metrics is **k = 19**.

Selecting **k = 19** ensures robust performance across all metrics while
avoiding overfitting at smaller k values or underfitting at larger ones.

```{r}
churn_col <- which(colnames(cust_scaled_train) == "Churn")
prediction_knn <- knn(train = cust_scaled_train[,-churn_col], 
                      test = cust_scaled_test[,-churn_col], 
                      cl = cust_scaled_train[,churn_col], 
                      k = 19)
summary(prediction_knn)

confusionMatrix(as.factor(prediction_knn), as.factor(cust_scaled_test$Churn), positive = "1")
```

The KNN model achieves an accuracy of 77.5%, which is better than the
Logistic Regression but worse than the SVM, indicating reasonable
predictive performance. The Kappa statistic (0.4113) reflects the worse
agreement beyond chance, although not being incredibly bad. However, the
sensitivity (55.63%) is relatively low if compared to the Logistic
Regression, meaning the model struggles to identify churners
effectively, which could limit its usefulness in retention strategies.
While the specificity (85.20%) indicates strong performance in
predicting non-churners, improving sensitivity is crucial to better
capture at-risk customers.

**Second Level Decision Tree:**

Now with all three models built, I will create a second-level (stack)
decision tree to combine their predictions, leveraging the strengths of
each model to improve overall accuracy by using their outputs as
features to make the final classification decision.

The first thing to do is to combine all predictions into a new
dataframe, as done with the code below:

```{r}
# Renaming the original target variable from the test set:
response <- cust_encoded_test$Churn

all_predictions <- data.frame(
                              binary_prediction_lr = as.numeric(binary_prediction_lr), 
                              prediction_knn = as.numeric(prediction_knn), 
                              svm_prediction = as.numeric(svm_prediction),
                              response = as.numeric(response))

str(all_predictions)

# Just a formality to make sure all positives are = 1
all_predictions$prediction_knn <- ifelse(all_predictions$prediction_knn == 1, 0, 1)
all_predictions$svm_prediction <- ifelse(all_predictions$svm_prediction == 1, 0, 1)
```

With the dataset create, and correctly formatted, we can split it to
apply the Decision Tree. That's what the code below does:

```{r}
test_rate <- 0.3
set.seed(1234567)

test_prediction_rows <- sample(1:nrow(all_predictions), test_rate*nrow(all_predictions))
train_predictions <- all_predictions[-test_prediction_rows, ]
test_predictions <- all_predictions[test_prediction_rows, ]
```

Once with the splits we apply a simple decision tree.

```{r}
model <- C5.0(as.factor(response) ~ ., data = train_predictions)
plot(model)

predictions <- predict(model, newdata = test_predictions)

confusionMatrix(as.factor(predictions),
                as.factor(test_predictions$response),
                positive="1")
```

From the metrics of this simple decision tree model we can see that
although the accuracy is really high, the sensitivity is considerably
low if compared to some of the individual models. To try to solve that
and better balance the accuracy/sensitivity trade-off, I will apply a
cost matrix that penalizes the false negatives, which will increase the
false positives, but likely, the true positives as well.

```{r}
cost_matrix_prediction <- matrix(c(0, 3,   
                                  1, 0),  
                                  nrow = 2,
                                  byrow = TRUE)
rownames(cost_matrix_prediction) <- colnames(cost_matrix_prediction) <- c(0, 1)

model2 <- C5.0(as.factor(response) ~ ., data = train_predictions, costs = cost_matrix_prediction)
plot(model2)
predictions2 <- predict(model2, newdata = test_predictions)

confusionMatrix(as.factor(predictions2),
                as.factor(test_predictions$response),
                positive="1")
```

This final stacked model focuses on maximizing sensitivity, achieving
the highest sensitivity (0.7665 with a 3:1 cost matrix) but sacrifices
some of the original accuracy (0.7609) and Kappa (0.4477), indicating
less balanced predictions. In contrast, the SVM achieves the highest
accuracy (0.7981) and a moderate Kappa (0.4551) but has poor sensitivity
(0.5541), struggling to identify positive churn cases. Logistic
regression strikes a good balance, with accuracy (0.7582), sensitivity
(0.7530), and Kappa (0.4497), making it more consistent across metrics.
KNN falls behind with slightly lower accuracy (0.775) and sensitivity
(0.5563).

In general, the stack model makes a good job at being consistent across
the metrics.

------------------------------------------------------------------------

## Part 2: Implementing the Prediction Model

Whenever trying to implement a model, the first step is to simply set
the key definitions:

-   Cost of discount per customer: \$100

-   Retention rate due to discount: 30% of customers predicted to churn
    who were actually going to churn will stay.

-   Savings per retained customer: \$500 (avoided customer acquisition
    cost).

Let's make those numbers variables, so I can make some basic
calculations with them:

```{r}
cost_discount_per_customer <- 100
retention_rate <- 0.3
savings_per_retained_customer <- 500

true_positives <- 229
false_positives <-  170

cost_of_discount <- (true_positives + false_positives) * cost_discount_per_customer
retention_savings <- (retention_rate * true_positives) * savings_per_retained_customer

net_financial_impact <- retention_savings - cost_of_discount

```

Currently, the overall cost of discount to be offered to all clients
predicted to leave is \$`r cost_of_discount`. That cost implies a
success of 30% of the ones that were going to leave (true positives)
were convinced, and that generates an overall saving of
\$`r retention_savings`. Based on these two information, we can conclude
that the net financial impact is of \$`r net_financial_impact`.

As seen, if the Customer Retention Incentive Program is made with the
current prediction model, the company will loose money. However, in
order to solve this problem, there is an important assumption to make.
We will be considering that the \$/500 savings of new acquisition will
only apply true positives that we are able to convince, but they will
not translate to the idea that each customer lost implies a cost of
\$/500.

What that in mind, I will change the weights on the cost matrix to
maximize the financial impact. As for now, the cost matrix is giving a
higher penalty for the false negatives (vs. the false positives),
nonetheless, with the cost of discount that the false positives bring to
the program, I will try to invert the weights, in order to penalize more
the false positives.

The code below changes the cost matrix and generates a third stacked
model:

```{r}
cost_matrix_prediction2 <- matrix(c(0, 1,   
                                  1, 0),  
                                  nrow = 2,
                                  byrow = TRUE)
rownames(cost_matrix_prediction2) <- colnames(cost_matrix_prediction2) <- c("0", "1")

model3 <- C5.0(as.factor(response) ~ ., data = train_predictions, costs = cost_matrix_prediction2)
plot(model3)
predictions3 <- predict(model3, newdata = test_predictions)

confusionMatrix(as.factor(predictions3),
                as.factor(test_predictions$response),
                positive="1")

new_true_positives <- 160
new_false_positives <- 64 

cost_of_discount2 <- (new_true_positives + new_false_positives) * cost_discount_per_customer
retention_savings2 <- (retention_rate * new_true_positives) * savings_per_retained_customer

net_financial_impact2 <- retention_savings2 - cost_of_discount2
```

The code below changes the cost matrix once more and generates a fourth
stacked model:

```{r}
cost_matrix_prediction3 <- matrix(c(0, 1,   
                                  1.5, 0),  
                                  nrow = 2,
                                  byrow = TRUE)
rownames(cost_matrix_prediction3) <- colnames(cost_matrix_prediction3) <- c("0", "1")

model4 <- C5.0(as.factor(response) ~ ., data = train_predictions, costs = cost_matrix_prediction3)
plot(model4)
predictions4 <- predict(model4, newdata = test_predictions)

confusionMatrix(as.factor(predictions4),
                as.factor(test_predictions$response),
                positive="1")

new2_true_positives <- 139
new2_false_positives <- 49 

cost_of_discount3 <- (new2_true_positives + new2_false_positives) * cost_discount_per_customer
retention_savings3 <- (retention_rate * new2_true_positives) * savings_per_retained_customer

net_financial_impact3 <- retention_savings3 - cost_of_discount3
```

------------------------------------------------------------------------

### Conclusion:

Although this final stacked model just generated deliver a profitable
scenario for the Customer Retention Incentive Program, and might be the
best one to be applied to this specific case, it does not mean it is the
better performing model for an overall prediction.

The conclusion here is that the best model you can generate will depend
of your objectives and goals for it. The model to make the more accurate
future predictions is likely the last one from part 1, but for this
business case, the model4 might be the best.
